\chapter*{Executive Summary}
\addcontentsline{toc}{chapter}{Executive Summary}

Training tokens are observations in a normalized autoregressive probability model. Their categorical negative log likelihood is cross-entropy. VFE 4.0 retains this observation term and augments it with the complexity of latent states, model states, source assignments, and, when geometry is random rather than fixed, geometric variables. The proposed construction is therefore a latent-variable language model whose observation semantics remain those of ordinary probabilistic language modeling.

The paper later fixes one normalized causal joint and one normalized structured variational law $Q$. Conditional on those definitions and their stated support and integrability assumptions, the evidence decomposition is an exact algebraic identity and its negative evidence lower bound is the VFE 4.0 objective. This identity does not establish that $Q$ equals the posterior, that a truncated or stochastic optimizer converges, or that the resulting model improves prediction.

VFE 4.0 is a research proposal, not a report of VFE 4.0 experimental results. Perplexity, calibration, posterior collapse, curvature effects, and computational scaling remain empirical questions. The purpose of the construction is to make those questions falsifiable under one probabilistic objective rather than to answer them in advance.

\chapter{Scope, Contribution, and Claim Status}

The proposed contribution is a gauge-covariant causal latent-variable language model on associated bundles whose state inference, model-channel inference, source selection, and categorical decoder learning are evaluated against one evidence lower bound. The generative model is defined before its recognition law and optimization procedure. Consequently, the later ELBO identity is conditional on the fixed normalized joint and normalized $Q$ that the paper specifies; it is not a premise from which those objects are inferred.

The manuscript separates mathematical statements from engineering decisions and empirical predictions according to \cref{tab:claim-status}. This separation applies throughout: a valid probability model can support exact identities even when posterior approximation, optimizer behavior, predictive quality, and computational feasibility remain unresolved.

\begin{sciencetable}{Claim status used throughout the VFE 4.0 white paper.}
\label{tab:claim-status}
\begin{tabularx}{\textwidth}{>{\raggedright\arraybackslash}p{0.19\textwidth} >{\raggedright\arraybackslash}X >{\raggedright\arraybackslash}p{0.25\textwidth}}
\toprule
\textbf{Status} & \textbf{Content} & \textbf{Required support} \\
\midrule
Definition & Causal token graph; principal and associated bundles; normalized transition and emission factors; source variables; structured recognition law; information-form Gaussian family; VFE 4.0 objective. & Explicit notation, domains, conditioning sets, and normalization assumptions. \\
\addlinespace
Identity/Lemma & Normalization of the directed joint; the ELBO evidence identity and its factorwise decomposition; Gaussian coordinate conversion and factor assembly; gauge transformation laws. & Algebra from the later-defined normalized joint and $Q$, with every assumption stated. \\
\addlinespace
Proposition & Gauge invariance under simultaneous pushforward; preservation of token-block precision sparsity; Regime I flatness and identity loop holonomy; nonclosure of diagonal covariance under general $\mathrm{GL}(K)$. & Proof or proof sketch with the applicable group, family, and regularity conditions. \\
\addlinespace
Implementation choice & Sparse Cholesky or related sparse factorization; solver, damping, storage, and approximation choices. & Numerical specification and verification against analytic reference problems. \\
\addlinespace
Empirical hypothesis & Predictive perplexity, calibration, posterior collapse, curvature effects, and scaling behavior, together with posterior-structure and optimization hypotheses. & Preregistered controls, uncertainty across runs, and explicit falsification criteria. \\
\bottomrule
\end{tabularx}
\end{sciencetable}

The exact core excludes claims that are not consequences of its factorization. The paper therefore does not call the recognition law exact, infer optimizer convergence from the ELBO identity, or treat gauge covariance as evidence of predictive benefit. Extensions such as latent connection variables, nonflat holonomy, configuration-space Gibbs laws, and additional multihop factors require their own normalized constructions before they can enter the same claim class.
