\chapter{Structured Recognition in Information Coordinates}

This chapter defines normalized recognition laws over the complete latent inventory of \cref{eq:vfe4-generative-joint} and then identifies the conditional Gaussian subfamily used by the reference implementation.  The distinction between filtering, smoothing, and prediction is part of the probability model: the first two are observation-conditioned variational laws, whereas prediction uses the target-blind causal prior.

\section{Normalized filtering and smoothing laws}

For $t\geq1$, define the recognition history
\begin{equation}
\mathcal H_t
:=(z_{0:t-1},m_{0:t-1},a_{1:t-1},b_{1:t-1}),
\label{eq:recognition-history}
\end{equation}
and set $x_{\leq0}:=\varnothing$.  The information available to the two recognition regimes is
\begin{equation}
\mathcal I_t^{\mathrm f}:=x_{\leq t},
\qquad
\mathcal I_t^{\mathrm s}:=x_{1:T},
\qquad t=0,\ldots,T.
\label{eq:filtering-smoothing-information}
\end{equation}
Thus the filtering factor at position $t$ may use the current observation $x_t$ but no later token, while every smoothing factor may use the complete observed sequence.  For either regime $r\in\{\mathrm f,\mathrm s\}$, let
\begin{equation}
\begin{aligned}
Q_\psi^{(r)}(\boldsymbol z,\boldsymbol m,a,b\given x,\Gamma)
:={}&Q_{\psi,0}^{(r)}
(z_0,m_0\given\mathcal I_0^{(r)},\Gamma)\\
&\times\prod_{t=1}^{T}
Q_{\psi,t}^{(r)}
(z_t,m_t,a_t,b_t\given
\mathcal H_t,\mathcal I_t^{(r)},\Gamma).
\end{aligned}
\label{eq:normalized-filtering-smoothing-recognition}
\end{equation}
The source coordinates in each time-slice factor have exactly the supports declared by the generative joint,
\begin{equation}
a_t\in\operatorname{Pa}_z(t),
\qquad
b_t\in\operatorname{Pa}_m(t),
\label{eq:recognition-source-supports}
\end{equation}
and the factor is zero outside those sets.  In the local frames of \cref{eq:normalized-initial-law}, normalization means
\begin{equation}
\begin{aligned}
\int Q_{\psi,0}^{(r)}
\mathop{\mathrm d\lambda_{z,0}}
\mathop{\mathrm d\lambda_{m,0}}&=1,\\
\sum_{a_t\in\operatorname{Pa}_z(t)}
\sum_{b_t\in\operatorname{Pa}_m(t)}
\int Q_{\psi,t}^{(r)}
\mathop{\mathrm d\lambda_{z,t}}
\mathop{\mathrm d\lambda_{m,t}}&=1,
\end{aligned}
\label{eq:recognition-factor-normalization}
\end{equation}
for every fixed value of the displayed conditioning variables.  Repeated conditional integration and summation therefore normalize \cref{eq:normalized-filtering-smoothing-recognition}.  The joint time-slice notation imposes no within-slice mean-field assumption; a particular implementation may further factor a slice only after declaring the corresponding conditional independences.

Dependence on $\mathcal H_t$ allows a later latent block to depend on earlier latent blocks.  Consequently, even if the innovation within each conditional factor is block diagonal, the marginal path law can carry cross-token and cross-channel covariance.  Removing history dependence is a restriction of the recognition family, not a consequence of filtering.

The causal predictive law at position $t$ is a different object.  Before observing $x_t$, \cref{eq:vfe4-generative-joint} induces
\begin{equation}
\begin{aligned}
P_{\theta,t}^{\mathrm{pred}}
(z_t,m_t,a_t,b_t\given\mathcal H_t,x_{<t},\Gamma)
:={}&\pi_t^m(b_t)
K^m_{\theta,tb_t}(m_t\given m_{b_t},x_{<t},\Gamma)\\
&\times\pi_t^z(a_t)
K^z_{\theta,ta_t}(z_t\given z_{a_t},m_t,x_{<t},\Gamma).
\end{aligned}
\label{eq:target-blind-predictive-prior}
\end{equation}
This normalized prior contains neither $x_t$ nor a variational parameter.  Training may use $Q_\psi^{(\mathrm f)}$ or $Q_\psi^{(\mathrm s)}$ after observations are supplied, but generation and predictive evaluation must propagate \cref{eq:target-blind-predictive-prior} and the categorical emission rather than substituting either recognition law.

\section{Conditional joint Gaussian information form}

Stack the continuous path in token-major order as
\begin{equation}
y:=\operatorname{col}(z_0,m_0,\ldots,z_T,m_T)
\in\mathbb R^D,
\qquad
D=(T+1)(d_z+d_m).
\label{eq:stacked-continuous-path}
\end{equation}
Because $(a,b)$ are finite discrete variables, any recognition law admits the disintegration
\begin{equation}
Q_\psi^{(r)}(y,a,b\given x,\Gamma)
=w_\psi^{(r)}(a,b\given x,\Gamma)
q_\psi^{(r)}(y\given a,b,x,\Gamma).
\label{eq:recognition-disintegration}
\end{equation}
With counting measure on $(a,b)$ and Lebesgue measure on $y$, the entropy of the complete mixed recognition law is
\begin{equation}
\begin{aligned}
H\left(Q_\psi^{(r)}\right)
&:=-\sum_{a,b}\int
Q_\psi^{(r)}(y,a,b\given x,\Gamma)
\log Q_\psi^{(r)}(y,a,b\given x,\Gamma)
\mathop{\mathrm dy}\\
&=-\sum_{a,b}w_\psi^{(r)}(a,b\given x,\Gamma)
\int q_\psi^{(r)}(y\given a,b,x,\Gamma)
\left[
\log w_\psi^{(r)}(a,b\given x,\Gamma)
+\log q_\psi^{(r)}(y\given a,b,x,\Gamma)
\right]
\mathop{\mathrm dy}\\
&=H\left(w_\psi^{(r)}\right)
+\sum_{a,b}w_\psi^{(r)}(a,b\given x,\Gamma)
H\left(q_\psi^{(r)}(\mathord{\cdot}\given a,b,x,\Gamma)\right).
\end{aligned}
\label{eq:mixed-recognition-joint-entropy}
\end{equation}
Here $H(w):=-\sum_{a,b}w(a,b)\log w(a,b)$, with $0\log0:=0$.  The second equality uses $\log(wq)=\log w+\log q$ on every positive-weight component.  Equation~\eqref{eq:mixed-recognition-joint-entropy} is the joint entropy of the categorical sources and continuous path that enters the ELBO, so the source entropy cannot be omitted.

This joint mixed entropy is not the differential entropy of the source-marginal mixture in \cref{eq:source-marginal-gaussian-mixture}.  The latter is
\begin{equation}
H\left(Q_\psi^{(r)}(y\given x,\Gamma)\right)
=-\int
\left(\sum_{a,b}w_{ab}q_{ab}(y)\right)
\log\left(\sum_{a,b}w_{ab}q_{ab}(y)\right)
\mathop{\mathrm dy},
\label{eq:source-marginal-mixture-entropy}
\end{equation}
which is generally not the weighted component entropy $\sum_{a,b}w_{ab}H(q_{ab})$ and does not inherit the componentwise Gaussian closed form.

The reference Gaussian family is a restriction only on the continuous conditional in \cref{eq:recognition-disintegration}.  For each source assignment with positive weight, it is
\begin{equation}
q(y)
=\exp\left(
h^\top y-\frac12y^\top Jy-A(h,J)
\right),
\qquad
J=J^\top\succ0,
\label{eq:joint-gaussian-information-form}
\end{equation}
where dependence of $(h,J)$ on $(r,a,b,x,\Gamma)$ is suppressed and
\begin{equation}
A(h,J)
=\frac12h^\top J^{-1}h
-\frac12\log\det J
+\frac D2\log(2\pi).
\label{eq:gaussian-log-normalizer}
\end{equation}
Completion of the square, given in the appendix, yields
\begin{equation}
\mu=J^{-1}h,
\qquad
\Sigma=J^{-1},
\qquad
h=J\mu.
\label{eq:information-moment-conversion}
\end{equation}
These relations are solve interfaces: they do not require constructing a dense inverse.

Marginalizing the source variables gives
\begin{equation}
Q_\psi^{(r)}(y\given x,\Gamma)
=\sum_{a,b}w_\psi^{(r)}(a,b\given x,\Gamma)
\mathcal N(y;\mu_{ab},\Sigma_{ab}),
\label{eq:source-marginal-gaussian-mixture}
\end{equation}
which is generally a Gaussian mixture rather than one Gaussian.  If a later computation replaces this mixture by a single Gaussian, that replacement is a declared moment projection.  With $w_c$ denoting a source-assignment weight, the matching moments are
\begin{equation}
\begin{aligned}
\bar\mu&=\sum_cw_c\mu_c,\\
\bar\Sigma&=\sum_cw_c
\left[
\Sigma_c+(\mu_c-\bar\mu)(\mu_c-\bar\mu)^\top
\right].
\end{aligned}
\label{eq:mixture-moment-projection}
\end{equation}
Equation~\eqref{eq:mixture-moment-projection} defines an approximation to the mixture, not an identity of distributions.

\section{Natural, expectation, and moment coordinates}

Using the Frobenius pairing $\langle A,B\rangle_{\mathrm F}:=\operatorname{tr}(A^\top B)$, the sufficient statistics and their natural parameters are
\begin{equation}
\begin{aligned}
T_1(y)&=y,
&\eta_1&=h,\\
T_2(y)&=yy^\top,
&\eta_2&=-\frac12J.
\end{aligned}
\label{eq:gaussian-natural-coordinates}
\end{equation}
The exact expectation coordinates dual to \cref{eq:gaussian-natural-coordinates} are
\begin{equation}
\tau_1=\mathbb E_q[y]=\mu,
\qquad
\tau_2=\mathbb E_q[yy^\top]
=M=\Sigma+\mu\mu^\top.
\label{eq:gaussian-expectation-coordinates}
\end{equation}
Matrix differentiation on the symmetric domain gives
\begin{equation}
\nabla_hA=\mu,
\qquad
\nabla_JA=-\frac12M.
\label{eq:gaussian-log-normalizer-gradients}
\end{equation}
Equivalently, differentiating with respect to $\eta_2=-J/2$ gives $\nabla_{\eta_2}A=M$.  The pair $(\mu,\Sigma)$ remains the most useful moment view for uncertainty summaries, sampling, and emissions, but it is not the exact Fisher-dual pair to $(h,-J/2)$ because the second sufficient statistic is the uncentered moment $M$, not the covariance $\Sigma$.  This exponential-family distinction follows the standard natural and mean-parameter geometry \citep{Wainwright2008,Amari2016}.

The conditional mean-field control is the nested restriction
\begin{equation}
J_{\mathrm{MF}}
=\operatorname{diag}
(J_{z,0},J_{m,0},\ldots,J_{z,T},J_{m,T}),
\label{eq:mean-field-information-restriction}
\end{equation}
which makes the continuous Gaussian component factor across every state and model block.  Removing source dependence as well gives the fully factorized control
\begin{equation}
Q_{\mathrm{MF}}(\boldsymbol z,\boldsymbol m\given x,\Gamma)
=\prod_{t=0}^{T}
q_t^z(z_t\given x,\Gamma)
q_t^m(m_t\given x,\Gamma).
\label{eq:full-mean-field-control}
\end{equation}
Neither restriction is the default structured family.

\section{Joint entropy and oriented Gaussian KL}

The differential entropy of the complete conditional Gaussian is
\begin{equation}
\begin{aligned}
H(q)
&=-\mathbb E_q[\log q(y)]\\
&=\frac12
\left[
D\bigl(1+\log(2\pi)\bigr)-\log\det J
\right]\\
&=\frac12
\left[
D\bigl(1+\log(2\pi)\bigr)+\log\det\Sigma
\right].
\end{aligned}
\label{eq:full-joint-gaussian-entropy}
\end{equation}
For token blocks $y_t=(z_t,m_t)$, the difference
\begin{equation}
\sum_{t=0}^{T}H(q(y_t))-H(q(y))
=D_{\mathrm{KL}}
\left(
q(y)\middle\Vert\prod_{t=0}^{T}q(y_t)
\right)
\geq0,
\label{eq:gaussian-total-correlation}
\end{equation}
is the total correlation.  Replacing the joint entropy by a sum of node entropies therefore drops this dependence term and overcounts the entropy unless the token blocks are independent.

Let $q_i=\mathcal N(\mu_i,\Sigma_i)$ with $J_i=\Sigma_i^{-1}$ for $i\in\{0,1\}$.  With the orientation $D_{\mathrm{KL}}(q_0\Vert q_1)$ fixed, the covariance form is
\begin{equation}
\begin{aligned}
D_{\mathrm{KL}}(q_0\Vert q_1)
=\frac12\Bigl[
&\operatorname{tr}(\Sigma_1^{-1}\Sigma_0)
+(\mu_1-\mu_0)^\top
\Sigma_1^{-1}(\mu_1-\mu_0)\\
&-D+\log\frac{\det\Sigma_1}{\det\Sigma_0}
\Bigr],
\end{aligned}
\label{eq:gaussian-kl-covariance-form}
\end{equation}
and the identical precision form is
\begin{equation}
\begin{aligned}
D_{\mathrm{KL}}(q_0\Vert q_1)
=\frac12\Bigl[
&\operatorname{tr}(J_1J_0^{-1})
+(\mu_1-\mu_0)^\top J_1(\mu_1-\mu_0)\\
&-D+\log\frac{\det J_0}{\det J_1}
\Bigr].
\end{aligned}
\label{eq:gaussian-kl-precision-form}
\end{equation}
In natural and expectation coordinates, the same orientation is
\begin{equation}
\begin{aligned}
D_{\mathrm{KL}}(q_0\Vert q_1)
={}&A(h_1,J_1)-A(h_0,J_0)
-(h_1-h_0)^\top\mu_0\\
&+\frac12\langle J_1-J_0,M_0\rangle_{\mathrm F}.
\end{aligned}
\label{eq:gaussian-kl-natural-expectation-form}
\end{equation}
The appendix derives these identities and audits their determinant signs.  They are standard Gaussian and exponential-family identities \citep{Kullback1951,petersen2012matrix,Wainwright2008}.

\section{Global gauge action and family closure}

Using the fiber coordinate matrices in \cref{eq:fiber-coordinate-change-matrices}, define the combined token maps and their block-diagonal global assembly by
\begin{equation}
\mathsf G_t
:=\operatorname{diag}(\mathsf G_{z,t},\mathsf G_{m,t}),
\qquad
\mathsf G
:=\operatorname{diag}(\mathsf G_0,\ldots,\mathsf G_T).
\label{eq:block-diagonal-global-frame-map}
\end{equation}
The stacked path transforms as $y'=\mathsf G y$.  Pushforward of the Gaussian probability measure then gives
\begin{equation}
\begin{aligned}
h'&=\mathsf G^{-\top}h,
&J'&=\mathsf G^{-\top}J\mathsf G^{-1},\\
\mu'&=\mathsf G\mu,
&\Sigma'&=\mathsf G\Sigma\mathsf G^\top,\\
&&M'&=\mathsf G M\mathsf G^\top.
\end{aligned}
\label{eq:joint-gaussian-gauge-laws}
\end{equation}
The coordinate density and log normalizer obey
\begin{equation}
q'(y')=\frac{q(y)}{|\det\mathsf G|},
\qquad
A(h',J')=A(h,J)+\log|\det\mathsf G|.
\label{eq:joint-density-log-normalizer-jacobian}
\end{equation}
Accordingly,
\begin{equation}
H(q')=H(q)+\log|\det\mathsf G|,
\qquad
D_{\mathrm{KL}}(q_0'\Vert q_1')
=D_{\mathrm{KL}}(q_0\Vert q_1).
\label{eq:entropy-shift-kl-invariance}
\end{equation}
Differential entropy is coordinate dependent, while the common Jacobian cancels from the log-density ratio defining KL.

Partition $J$ into combined token blocks $J_{ts}$.  Because \cref{eq:block-diagonal-global-frame-map} does not mix token indices,
\begin{equation}
J_{ts}'
=\mathsf G_t^{-\top}J_{ts}\mathsf G_s^{-1}.
\label{eq:precision-token-block-gauge-law}
\end{equation}
Invertibility implies $J_{ts}'=0$ if and only if $J_{ts}=0$.  Token-block sparsity, including block-banded and block-tridiagonal patterns, is therefore closed under the local gauge action.

Within-token scalar diagonality is not closed under general $\mathrm{GL}(K)$.  For example,
\begin{equation}
\Sigma=I_2,
\qquad
G=\begin{pmatrix}1&1\\0&1\end{pmatrix},
\qquad
G\Sigma G^\top
=\begin{pmatrix}2&1\\1&1\end{pmatrix}.
\label{eq:diagonal-covariance-nonclosure-example}
\end{equation}
Thus a full-$\mathrm{GL}(K)$ path must retain full within-token blocks, choose another representation class closed under the action, or restrict the group.  Rewriting a nonclosed family in natural coordinates does not restore closure.

\section{Sparse precision as the scalable interface}

A sparse Markov precision need not have a sparse covariance.  The scalar tridiagonal example
\begin{equation}
J=\begin{pmatrix}
2&-1&0\\
-1&2&-1\\
0&-1&2
\end{pmatrix},
\qquad
J^{-1}=\frac14\begin{pmatrix}
3&2&1\\
2&4&2\\
1&2&3
\end{pmatrix},
\label{eq:tridiagonal-precision-dense-covariance}
\end{equation}
already has a nonzero covariance between its first and third nodes.  The scalable information-form interface is therefore a factorization of $J$ together with linear solves $Jv=r$, selected inverse blocks required by the objective, log-determinant evaluation, and sampling.  If $J=LL^\top$ is a sparse Cholesky factorization, then
\begin{equation}
\log\det J=2\sum_i\log L_{ii},
\qquad
y=\mu+L^{-\top}\epsilon,
\qquad
\epsilon\sim\mathcal N(0,I_D),
\label{eq:sparse-factor-logdet-sampling}
\end{equation}
provide the determinant and sample interfaces without a dense covariance allocation.

Positive definiteness is part of the family definition.  It may be preserved by assembling $J$ from normalized Gaussian factors or by parameterizing a valid sparse factorization, but an unconstrained entrywise update to $J$ is not an admissible default.  Ordinary Euclidean SGD or Adam on $(h,J)$ is likewise not a natural gradient merely because $(h,-J/2)$ are natural coordinates; that name requires the Fisher metric or the exact natural--expectation dual update developed in the next chapter.
