\chapter{Information-Form Identities}

This appendix records supporting coordinate and factor-assembly identities that are used by the main construction. It follows the main text so that these details support, rather than precede, the definitions and ELBO dependency order.

\section{Gaussian completion, normalization, and dual derivatives}

Fix $h\in\mathbb R^D$ and $J=J^\top\succ0$, and define $\mu:=J^{-1}h$.  Direct expansion gives
\begin{equation}
\begin{aligned}
h^\top y-\frac12y^\top Jy
&=-\frac12
\left(y-J^{-1}h\right)^\top
J
\left(y-J^{-1}h\right)
+\frac12h^\top J^{-1}h\\
&=-\frac12(y-\mu)^\top J(y-\mu)
+\frac12h^\top J^{-1}h.
\end{aligned}
\label{eq:information-form-completion-square}
\end{equation}
The positive-definite Gaussian integral is
\begin{equation}
\int_{\mathbb R^D}
\exp\left[-\frac12(y-\mu)^\top J(y-\mu)\right]
\mathop{\mathrm dy}
=(2\pi)^{D/2}(\det J)^{-1/2}.
\label{eq:precision-gaussian-integral}
\end{equation}
Combining \cref{eq:information-form-completion-square,eq:precision-gaussian-integral} yields
\begin{equation}
\log\int_{\mathbb R^D}
\exp\left(h^\top y-\frac12y^\top Jy\right)
\mathop{\mathrm dy}
=\frac12h^\top J^{-1}h
-\frac12\log\det J
+\frac D2\log(2\pi),
\label{eq:derived-gaussian-log-normalizer}
\end{equation}
which proves \cref{eq:gaussian-log-normalizer}.  The completed density is $\mathcal N(\mu,J^{-1})$, establishing \cref{eq:information-moment-conversion}.

For the derivative calculation, set $\Sigma:=J^{-1}$ and $M:=\Sigma+\mu\mu^\top$.  The inverse and determinant differentials are
\begin{equation}
\mathop{\mathrm d}(J^{-1})
=-J^{-1}(\mathop{\mathrm d}J)J^{-1},
\qquad
\mathop{\mathrm d}\log\det J
=\operatorname{tr}(J^{-1}\mathop{\mathrm d}J).
\label{eq:inverse-logdet-differentials}
\end{equation}
Holding the other argument fixed for each partial differential gives
\begin{equation}
\begin{aligned}
\mathop{\mathrm d}_hA
&=\mu^\top\mathop{\mathrm d}h,\\
\mathop{\mathrm d}_JA
&=-\frac12\mu^\top(\mathop{\mathrm d}J)\mu
-\frac12\operatorname{tr}(\Sigma\mathop{\mathrm d}J)\\
&=-\frac12\langle M,\mathop{\mathrm d}J\rangle_{\mathrm F}.
\end{aligned}
\label{eq:log-normalizer-differential-proof}
\end{equation}
Because $M$ and $J$ are symmetric, the Frobenius gradients are exactly $\nabla_hA=\mu$ and $\nabla_JA=-M/2$.  Under the linear reparameterization $\eta_2=-J/2$, the chain rule gives $\nabla_{\eta_2}A=M$, confirming that $(\mu,M)$, rather than $(\mu,\Sigma)$, is the expectation-coordinate partner of $(h,-J/2)$.

\section{Joint entropy and Gaussian KL derivations}

Taking expectation in \cref{eq:joint-gaussian-information-form} and using
\begin{equation}
\mathbb E_q[y^\top Jy]
=\operatorname{tr}(JM)
=D+\mu^\top J\mu,
\qquad
h^\top\mu=\mu^\top J\mu,
\label{eq:gaussian-quadratic-expectation}
\end{equation}
gives
\begin{equation}
\begin{aligned}
H(q)
&=A(h,J)-h^\top\mu
+\frac12\operatorname{tr}(JM)\\
&=\frac12
\left[D\bigl(1+\log(2\pi)\bigr)-\log\det J\right],
\end{aligned}
\label{eq:joint-entropy-derivation}
\end{equation}
including the full joint determinant.  If $q_t$ are the token marginals, then
\begin{equation}
\begin{aligned}
D_{\mathrm{KL}}
\left(q(y)\middle\Vert\prod_tq_t(y_t)\right)
&=\mathbb E_q\left[
\log q(y)-\sum_t\log q_t(y_t)
\right]\\
&=\sum_tH(q_t)-H(q),
\end{aligned}
\label{eq:total-correlation-entropy-proof}
\end{equation}
which proves the sign in \cref{eq:gaussian-total-correlation}.

For $q_i=\mathcal N(\mu_i,\Sigma_i)$, write $\delta:=\mu_0-\mu_1$.  The two quadratic expectations needed for the oriented ratio $\log(q_0/q_1)$ are
\begin{equation}
\begin{aligned}
\mathbb E_{q_0}
[(y-\mu_0)^\top\Sigma_0^{-1}(y-\mu_0)]
&=D,\\
\mathbb E_{q_0}
[(y-\mu_1)^\top\Sigma_1^{-1}(y-\mu_1)]
&=\operatorname{tr}(\Sigma_1^{-1}\Sigma_0)
+\delta^\top\Sigma_1^{-1}\delta.
\end{aligned}
\label{eq:gaussian-kl-quadratic-expectations}
\end{equation}
Substitution into $\mathbb E_{q_0}[\log q_0-\log q_1]$ proves \cref{eq:gaussian-kl-covariance-form}.  Replacing $\Sigma_i$ by $J_i^{-1}$ and using
\begin{equation}
\log\frac{\det\Sigma_1}{\det\Sigma_0}
=\log\frac{\det J_0}{\det J_1},
\label{eq:gaussian-kl-determinant-conversion}
\end{equation}
proves \cref{eq:gaussian-kl-precision-form} with the same orientation.  Finally, the exponential-family log ratio gives
\begin{equation}
\begin{aligned}
D_{\mathrm{KL}}(q_0\Vert q_1)
={}&A_1-A_0
-(h_1-h_0)^\top\mu_0\\
&+\frac12\langle J_1-J_0,M_0\rangle_{\mathrm F},
\end{aligned}
\label{eq:natural-expectation-kl-derivation}
\end{equation}
which is \cref{eq:gaussian-kl-natural-expectation-form}.

\section{Gauge pushforward and precision-family closure}

Let $y'=\mathsf G y$ with the block-diagonal invertible map in \cref{eq:block-diagonal-global-frame-map}.  Substituting $y=\mathsf G^{-1}y'$ into the exponent gives
\begin{equation}
h^\top y-\frac12y^\top Jy
=(\mathsf G^{-\top}h)^\top y'
-\frac12y'^\top
(\mathsf G^{-\top}J\mathsf G^{-1})y',
\label{eq:gauge-transformed-information-exponent}
\end{equation}
which proves the $(h,J)$ transformations in \cref{eq:joint-gaussian-gauge-laws}.  Inversion and expectation then give
\begin{equation}
\begin{aligned}
(J')^{-1}
&=\mathsf GJ^{-1}\mathsf G^\top,\\
\mu'=(J')^{-1}h'
&=\mathsf G\mu,\\
M'=\mathbb E[y'y'^\top]
&=\mathsf GM\mathsf G^\top.
\end{aligned}
\label{eq:gauge-transformed-moment-proof}
\end{equation}
Moreover,
\begin{equation}
\log\det J'
=\log\det J-2\log|\det\mathsf G|,
\label{eq:gauge-precision-logdet-shift}
\end{equation}
while the quadratic term $h^\top J^{-1}h$ is unchanged.  Hence $A'=A+\log|\det\mathsf G|$ and $H'=H+\log|\det\mathsf G|$.  The pushed-forward density acquires the inverse Jacobian, so both factors in a KL ratio acquire the same additive log-density shift.  Their difference is unchanged, proving \cref{eq:entropy-shift-kl-invariance}.

\paragraph{Proposition (closure of token-block precision sparsity).}
Let $\mathcal S$ be any set of token-index pairs and let
\begin{equation}
\mathcal J_{\mathcal S}
:=\{J=J^\top\succ0:J_{ts}=0\text{ for }(t,s)\notin\mathcal S\}.
\label{eq:token-block-sparse-precision-family}
\end{equation}
For every block-diagonal $\mathsf G=\operatorname{diag}(\mathsf G_0,\ldots,\mathsf G_T)$ with invertible blocks, the congruence $J'=\mathsf G^{-\top}J\mathsf G^{-1}$ maps $\mathcal J_{\mathcal S}$ onto itself.

\emph{Proof.}
Block multiplication gives $J_{ts}'=\mathsf G_t^{-\top}J_{ts}\mathsf G_s^{-1}$.  Thus $J_{ts}=0$ implies $J_{ts}'=0$.  Applying the same argument to $J=\mathsf G^\top J'\mathsf G$ proves the converse.  Congruence by an invertible matrix preserves positive definiteness, so every transformed matrix remains in the precision cone.  \hfill$\square$

The conclusion does not extend to elementwise diagonality within a token block.  The matrix in \cref{eq:diagonal-covariance-nonclosure-example} has determinant one, yet its congruence image of $I_2$ has off-diagonal entry one.  General $\mathrm{GL}(2)$ therefore takes the diagonal-covariance family outside itself.  The same example applies to a diagonal precision by replacing the frame map with its inverse transpose.

Finally, direct multiplication verifies that the two matrices in \cref{eq:tridiagonal-precision-dense-covariance} are inverses.  Hence even the shortest nondegenerate Markov chain can have a tridiagonal precision and a dense covariance.  Sparse storage promises exact factorization and solve interfaces for $J$; it does not promise sparse storage for $J^{-1}$.

\chapter{Proof Details and Verification Oracles}

This appendix collects proof details and reference checks for claims introduced in the main text. The derivations below state the corresponding assumptions and numerical oracles explicitly.

\section{Full normalization proof for the fixed conditional joint}

By \cref{eq:nonempty-causal-parent-sets}, every state or model source index lies on a permitted edge $(j,t)\in E_T$.  The transports $\Omega^z_{tj}$ and $\Omega^m_{tj}$ used by the corresponding Gaussian kernels are therefore defined by \cref{eq:typed-associated-transports}.

Let $\#_A$ denote counting measure on a finite set $A$.  In the local frames used to express the densities, the reference measure for the fixed-horizon variables $(x,\boldsymbol z,\boldsymbol m,a,b)$ is
\begin{equation}
\begin{aligned}
\mu_T
:={}&\left(\bigotimes_{t=1}^{T}\#_{\mathcal X}\right)
\otimes\left(\bigotimes_{t=0}^{T}
(\lambda_{z,t}\otimes\lambda_{m,t})\right)\\
&\otimes\left(\bigotimes_{t=1}^{T}
(\#_{\operatorname{Pa}_z(t)}\otimes
\#_{\operatorname{Pa}_m(t)})\right).
\end{aligned}
\label{eq:full-joint-reference-measure}
\end{equation}
Each factor is measurable by the assumptions preceding \cref{eq:vfe4-generative-joint}.  The finite products are therefore measurable, and the density is nonnegative, so Tonelli's theorem permits every integration and summation below without a prior integrability assumption.

For fixed earlier variables and fixed $\Gamma$, collect the factors at time $t$ into
\begin{equation}
\begin{aligned}
F_t
:={}&\pi_t^m(b_t)
K^m_{\theta,tb_t}(m_t\given m_{b_t},x_{<t},\Gamma)
\pi_t^z(a_t)\\
&\times K^z_{\theta,ta_t}(z_t\given z_{a_t},m_t,x_{<t},\Gamma)
L_{\theta,t}(x_t\given z_t,m_t,\Gamma).
\end{aligned}
\label{eq:time-slice-factor-for-normalization}
\end{equation}
Nested normalization gives the pointwise identity
\begin{equation}
\begin{aligned}
&\sum_{b_t\in\operatorname{Pa}_m(t)}\pi_t^m(b_t)
\int_{E_{m,t}}K^m_{\theta,tb_t}
\sum_{a_t\in\operatorname{Pa}_z(t)}\pi_t^z(a_t)
\int_{E_{z,t}}K^z_{\theta,ta_t}
\sum_{x_t\in\mathcal X}L_{\theta,t}(x_t\given z_t,m_t,\Gamma)
\mathop{\mathrm d\lambda_{z,t}(z_t)}
\mathop{\mathrm d\lambda_{m,t}(m_t)}\\
&\quad=
\sum_{b_t\in\operatorname{Pa}_m(t)}\pi_t^m(b_t)
\int_{E_{m,t}}K^m_{\theta,tb_t}
\sum_{a_t\in\operatorname{Pa}_z(t)}\pi_t^z(a_t)
\mathop{\mathrm d\lambda_{m,t}(m_t)}\\
&\quad=
\sum_{b_t\in\operatorname{Pa}_m(t)}\pi_t^m(b_t)
\int_{E_{m,t}}K^m_{\theta,tb_t}
\mathop{\mathrm d\lambda_{m,t}(m_t)}
=1.
\end{aligned}
\label{eq:time-slice-nested-normalization}
\end{equation}
The unshown conditioning arguments of the two transition kernels are exactly those in \cref{eq:time-slice-factor-for-normalization}.  In the first equality, categorical emission normalization removes the innermost sum, state-kernel normalization removes the $z_t$ integral, and \cref{eq:normalized-causal-source-priors} removes the $a_t$ sum.  In the second equality, model-kernel normalization removes the $m_t$ integral and the same source-prior identity removes the $b_t$ sum.

To apply \cref{eq:time-slice-nested-normalization} to the complete density, begin at the terminal vertex $T$.  There is no factor at an earlier time that depends on $x_T$, $z_T$, $m_T$, $a_T$, or $b_T$, because every parent index is strictly smaller than its receiver and every history is $x_{<t}$.  Factors at later times may depend on terminal-block variables at an intermediate stage, but reverse topological order removes all such factors before that stage is reached.  Hence the nested elimination at $T$ is pointwise in every earlier variable and replaces $F_T$ by one.

After eliminating the terminal block, the same argument applies to $F_{T-1}$.  Induction removes $F_t$ for $t=T,T-1,\ldots,1$.  History dependence is harmless throughout: when the time-$t$ identity is evaluated, $x_{<t}$ is held fixed, and all factors that could depend on $x_t$ have already been eliminated.  What remains is
\begin{equation}
\int_{E_{z,0}\times E_{m,0}}
p_{\theta,0}(z_0,m_0\given\Gamma)
\mathop{\mathrm d\lambda_{z,0}(z_0)}
\mathop{\mathrm d\lambda_{m,0}(m_0)}
=1,
\label{eq:normalization-proof-initial-remainder}
\end{equation}
by \cref{eq:normalized-initial-law}.  Thus the density over $(x,\boldsymbol z,\boldsymbol m,a,b)$ in \cref{eq:vfe4-generative-joint} defines a probability measure with respect to \cref{eq:full-joint-reference-measure} for every fixed deterministic $\Gamma$.  Under a local frame change, \cref{eq:receiver-density-kernel-measure-invariance,eq:initial-density-gauge-jacobian} preserve this probability measure even though its coordinate density and Lebesgue representatives change.  \hfill$\square$

\section{Linear-Gaussian information-factor assembly}

Consider one normalized transition factor
\begin{equation}
y_t=Ay_j+c+\epsilon,
\qquad
\epsilon\sim\mathcal N(0,R),
\qquad
P:=R^{-1}=P^\top\succ0,
\label{eq:linear-gaussian-factor-setup}
\end{equation}
and define its residual by
\begin{equation}
r:=y_t-Ay_j-c.
\label{eq:linear-gaussian-transition-residual}
\end{equation}
Direct expansion gives
\begin{equation}
\begin{aligned}
r^\top P r
={}&y_j^\top A^\top P A y_j
-y_j^\top A^\top P y_t
-y_t^\top P A y_j
+y_t^\top P y_t\\
&+2y_j^\top A^\top P c
-2y_t^\top P c
+c^\top P c.
\end{aligned}
\label{eq:expanded-linear-gaussian-residual}
\end{equation}
Since $P$ is symmetric, the two cross terms in the first line are transposes of the same scalar. Writing the log factor in canonical information form,
\begin{equation}
\log K(y_t\given y_j)
=
\begin{pmatrix}h_j\\h_t\end{pmatrix}^{\top}
\begin{pmatrix}y_j\\y_t\end{pmatrix}
-\frac12
\begin{pmatrix}y_j\\y_t\end{pmatrix}^{\top}
\begin{pmatrix}
J_{jj}&J_{jt}\\
J_{tj}&J_{tt}
\end{pmatrix}
\begin{pmatrix}y_j\\y_t\end{pmatrix}
+C_K,
\label{eq:linear-gaussian-canonical-factor}
\end{equation}
comparison with $-r^\top P r/2$ gives exactly
\begin{equation}
\boxed{
\begin{aligned}
J_{jj}&=A^\top P A,
&J_{jt}&=-A^\top P,\\
J_{tj}&=-P A,
&J_{tt}&=P,\\
h_j&=-A^\top P c,
&h_t&=P c.
\end{aligned}}
\label{eq:linear-gaussian-information-blocks}
\end{equation}
The remaining constant is
\begin{equation}
C_K
=-\frac12c^\top P c
-\frac{d_t}{2}\log(2\pi)
-\frac12\log\det R,
\label{eq:linear-gaussian-factor-constant}
\end{equation}
where $d_t:=\dim y_t$. Thus normalized Gaussian factors contribute additively to a global $(h,J)$, including their normalizing constants in log-density or ELBO evaluation. The identities in \cref{eq:linear-gaussian-information-blocks} do not require $A$ to be square.

\section{Analytic evidence and ELBO positive control}

An implementation-level positive control can use the one-step normalized linear-Gaussian model
\begin{equation}
\begin{aligned}
y_0&\sim\mathcal N(\mu_0,S_0),\\
y_1\given y_0&\sim\mathcal N(Ay_0+c,R),\\
x\given y_1&\sim\mathcal N(Cy_1+d,S).
\end{aligned}
\label{eq:linear-gaussian-positive-control-model}
\end{equation}
The Gaussian observation in this oracle is a tractable verification device, not a replacement for the categorical observation law in the exact language model. Marginalization gives
\begin{equation}
\begin{aligned}
\bar x
&=C(A\mu_0+c)+d,\\
S_x
&=C(AS_0A^\top+R)C^\top+S,\\
\log p(x)
&=-\frac12\left[
d_x\log(2\pi)+\log\det S_x
+(x-\bar x)^\top S_x^{-1}(x-\bar x)
\right].
\end{aligned}
\label{eq:linear-gaussian-positive-control-evidence}
\end{equation}
The posterior information matrix and vector are assembled from the initial law, \cref{eq:linear-gaussian-information-blocks}, and the fixed observation factor. The observation contributes
\begin{equation}
J_{11}\mathrel{+}=C^\top S^{-1}C,
\qquad
h_1\mathrel{+}=C^\top S^{-1}(x-d).
\label{eq:gaussian-observation-information-contribution}
\end{equation}
Solving $J_*\mu_*=h_*$ and $J_*\Sigma_*=I$ gives the analytic posterior. For any nonposterior Gaussian $Q=\mathcal N(\mu_Q,\Sigma_Q)$ with rational positive-definite parameters, the control evaluates
\begin{equation}
\mathcal L(Q)
=\mathbb E_Q[\log p(x,y_0,y_1)-\log Q(y_0,y_1)],
\label{eq:linear-gaussian-positive-control-elbo}
\end{equation}
from complete quadratic expectations and verifies
\begin{equation}
\log p(x)-\mathcal L(Q)
-\KL\left(Q\middle\Vert
\mathcal N(\mu_*,\Sigma_*)\right)=0.
\label{eq:linear-gaussian-positive-control-identity}
\end{equation}
The protocol first checks the six blocks in \cref{eq:linear-gaussian-information-blocks} by exact Hessian and gradient extraction, then checks \cref{eq:linear-gaussian-positive-control-identity} with exact rational arithmetic. A floating-point implementation must additionally compare monolithic log-ratio and decomposed-factor evaluations at the same $Q$ before any optimizer is introduced.

\section{Bundle-map dimension and type oracle}

The coordinate audit for \cref{eq:typed-associated-transports,eq:typed-model-state-morphism} begins with column vectors $z_j,z_t\in\mathbb R^{d_z}$ and $m_j,m_t\in\mathbb R^{d_m}$.  The represented state link has shape $d_z\times d_z$, the represented model link has shape $d_m\times d_m$, and the cross-channel morphism has shape $d_z\times d_m$.  Consequently,
\begin{equation}
\begin{aligned}
\underbrace{\Omega^z_{tj}}_{d_z\times d_z}
\underbrace{z_j}_{d_z\times 1}
&\in\mathbb R^{d_z},
&
\underbrace{\Omega^m_{tj}}_{d_m\times d_m}
\underbrace{m_j}_{d_m\times 1}
&\in\mathbb R^{d_m},\\
\underbrace{B_t}_{d_z\times d_m}
\underbrace{m_t}_{d_m\times 1}
&\in\mathbb R^{d_z}.
&&
\end{aligned}
\label{eq:bundle-map-dimension-oracle}
\end{equation}
The first and third outputs in \cref{eq:bundle-map-dimension-oracle} inhabit the same target fiber $E_{z,t}$ and may enter one state-kernel location parameter.  The model-link output inhabits $E_{m,t}$ and may enter a model-kernel location parameter.  Combining the latter directly with a state-fiber vector would be ill typed.

The gauge-law audit is equally mechanical.  With $\rho_z(g_t)\in\mathbb R^{d_z\times d_z}$ and $\rho_m(g_t)\in\mathbb R^{d_m\times d_m}$, every product in \cref{eq:link-morphism-gauge-laws} has the same shape as the unprimed map:
\begin{equation}
\begin{aligned}
(d_z\times d_z)(d_z\times d_z)(d_z\times d_z)&=d_z\times d_z,\\
(d_m\times d_m)(d_m\times d_m)(d_m\times d_m)&=d_m\times d_m,\\
(d_z\times d_z)(d_z\times d_m)(d_m\times d_m)&=d_z\times d_m.
\end{aligned}
\label{eq:gauge-law-dimension-oracle}
\end{equation}
No equality $d_z=d_m$ is used.  This check is the source-level oracle for later probability kernels.  A covariance on the state fiber is a positive symmetric contravariant tensor and its precision is a positive symmetric covariant tensor,
\begin{equation}
\Sigma_{z,t}\in\operatorname{Sym}^{2}_{+}(E_{z,t})
\simeq\operatorname{Hom}_{+}(E_{z,t}^{*},E_{z,t}),
\qquad
J_{z,t}\in\operatorname{Sym}^{2}_{+}(E_{z,t}^{*})
\simeq\operatorname{Hom}_{+}(E_{z,t},E_{z,t}^{*}).
\label{eq:state-covariance-precision-types}
\end{equation}
The model-fiber tensors have the analogous types with $E_{z,t}$ replaced by $E_{m,t}$.  If
\begin{equation}
\mathsf G_{z,t}:=\rho_z(g_t),
\qquad
\mathsf G_{m,t}:=\rho_m(g_t),
\label{eq:fiber-coordinate-change-matrices}
\end{equation}
then covariance and precision carry opposite congruence laws in either channel:
\begin{equation}
\begin{aligned}
\Sigma_{z,t}'&=\mathsf G_{z,t}\Sigma_{z,t}\mathsf G_{z,t}^{\top},
&J_{z,t}'&=\mathsf G_{z,t}^{-\top}J_{z,t}\mathsf G_{z,t}^{-1},\\
\Sigma_{m,t}'&=\mathsf G_{m,t}\Sigma_{m,t}\mathsf G_{m,t}^{\top},
&J_{m,t}'&=\mathsf G_{m,t}^{-\top}J_{m,t}\mathsf G_{m,t}^{-1}.
\end{aligned}
\label{eq:covariance-precision-opposite-congruence}
\end{equation}
A kernel over $E_{z,t}$ must therefore receive a location in $E_{z,t}$ together with a covariance in $\operatorname{Sym}^{2}_{+}(E_{z,t})$ or a precision in $\operatorname{Sym}^{2}_{+}(E_{z,t}^{*})$.  A kernel over $E_{m,t}$ must receive the corresponding model-fiber objects.

\section{Expanded Regime I composition check}

For a path $v_0\to v_1\to\cdots\to v_r$ in the auxiliary geometric graph, the composite group transport is ordered so that the rightmost map acts first.  Repeated use of \cref{eq:regime-i-cocycle-identity} gives
\begin{equation}
\Omega_{v_rv_{r-1}}\cdots\Omega_{v_1v_0}
=U_{v_r}U_{v_0}^{-1}.
\label{eq:regime-i-open-path-telescope}
\end{equation}
Thus an open-path transport depends only on its endpoints in Regime I.  Setting $v_r=v_0$ yields the identity, as in \cref{eq:regime-i-loop-holonomy}.  Applying either representation to \cref{eq:regime-i-open-path-telescope} gives the correctly typed composites
\begin{equation}
\Omega^z_{v_rv_{r-1}}\circ\cdots\circ\Omega^z_{v_1v_0}
:E_{z,v_0}\longrightarrow E_{z,v_r},
\qquad
\Omega^m_{v_rv_{r-1}}\circ\cdots\circ\Omega^m_{v_1v_0}
:E_{m,v_0}\longrightarrow E_{m,v_r}.
\label{eq:associated-open-path-types}
\end{equation}
This check uses a geometric walk in the underlying undirected graph or cell complex.  It does not insert a directed cycle into the causal factor graph.

\chapter{Notation}

The symbols in \cref{tab:vfe4-notation} fix the notation path used throughout the construction. The table records exact mappings among coordinate views but introduces no model assumption beyond the definitions in the main text.

{\small
\begin{longtable}{>{\raggedright\arraybackslash}p{0.18\textwidth} >{\raggedright\arraybackslash}p{0.27\textwidth} >{\raggedright\arraybackslash}p{0.45\textwidth}}
\caption{Notation for the gauge-causal VFE 4.0 construction.}
\label{tab:vfe4-notation}\\
\toprule
\textbf{Symbol} & \textbf{Domain or type} & \textbf{Meaning} \\
\midrule
\endfirsthead
\multicolumn{3}{c}{\tablename\ \thetable\ continued}\\
\toprule
\textbf{Symbol} & \textbf{Domain or type} & \textbf{Meaning} \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{Continued on the next page}\\
\endfoot
\bottomrule
\endlastfoot
$t,T$ & $t\in\{0,\ldots,T\}$ & Token position and fixed sequence horizon. \\
\addlinespace
$x_t,x_{<t}$ & $x_t\in\mathcal X$ & Observed token and realized causal prefix; $x_t$ is never a latent belief coordinate. \\
\addlinespace
$D_T=(V_T,E_T)$ & Directed acyclic graph & Causal token base with $(j,t)\in E_T$ only when $j<t$. \\
\addlinespace
$G,P$ & Matrix Lie group; principal $G$-bundle & Declared gauge group and the working product bundle $P=|D_T|\times G$. \\
\addlinespace
$E_z,E_m$ & Associated vector bundles & State and model bundles with fibers $E_{z,t}$ and $E_{m,t}$ of dimensions $d_z$ and $d_m$. \\
\addlinespace
$\rho_z,\rho_m$ & Representations of $G$ & Possibly distinct actions of the shared group on the state and model fibers. \\
\addlinespace
$z_t,m_t$ & $z_t\in E_{z,t}$, $m_t\in E_{m,t}$ & State and model latent variables; bold symbols $\boldsymbol z,\boldsymbol m$ denote their complete paths. \\
\addlinespace
$\Gamma$ & Deterministic geometric data in the exact core & Collection of declared frames, links, representations, morphisms, and other fixed geometric coefficients. \\
\addlinespace
$\Omega_{tj}$ & $G$-valued link & Group transport from source $j$ to receiver $t$; $\Omega^z_{tj}$ and $\Omega^m_{tj}$ are its represented state and model maps. \\
\addlinespace
$B_t$ & $\operatorname{Hom}(E_{m,t},E_{z,t})$ & Typed model-to-state morphism; no equality $d_m=d_z$ is assumed. \\
\addlinespace
$a_t,b_t$ & Finite categorical variables & State-source and model-source assignments on $\operatorname{Pa}_z(t)$ and $\operatorname{Pa}_m(t)$. \\
\addlinespace
$\pi_t^z,\pi_t^m$ & Normalized probability mass functions & Fixed generative priors for $a_t$ and $b_t$. \\
\addlinespace
$K^z_{\theta,tj},K^m_{\theta,tj}$ & Normalized Markov kernels & State and model transition densities for source $j$ and receiver $t$. \\
\addlinespace
$L_{\theta,t}$ & Categorical Markov kernel on $\mathcal X$ & Normalized token-emission law whose expected log probability is the observation term. \\
\addlinespace
$Y$ & $(\boldsymbol z,\boldsymbol m,a,b)$ & Complete latent inventory used in the evidence identity. \\
\addlinespace
$Q_\psi^{(r)}$ & Normalized mixed variational law & Filtering ($r=\mathrm f$) or smoothing ($r=\mathrm s$) recognition law over $Y$. \\
\addlinespace
$\beta_t,\gamma_t$ & Conditional categorical recognition laws & State-source and model-source routing factors; these are not the fixed priors $\pi_t^z,\pi_t^m$. \\
\addlinespace
$y$ & $\operatorname{col}(z_0,m_0,\ldots,z_T,m_T)\in\mathbb R^D$ & Token-major stacking of the continuous state and model paths in chosen local frames. \\
\addlinespace
$D$ & $(T+1)(d_z+d_m)$ & Dimension of the stacked continuous path $y$. \\
\addlinespace
$w(a,b),q(y\given a,b)$ & Categorical mass and conditional density & Disintegration of $Q(y,a,b)=w(a,b)q(y\given a,b)$; source marginalization generally gives a Gaussian mixture. \\
\addlinespace
$(h,J)$ & $h\in\mathbb R^D$, $J=J^\top\succ0$ & Information vector and precision; the Gaussian natural coordinates are $(h,-J/2)$. \\
\addlinespace
$(\mu,M)$ & $M=\mathbb E[yy^\top]$ & Expectation-coordinate pair exactly dual to $(h,-J/2)$. \\
\addlinespace
$(\mu,\Sigma)$ & $\Sigma=J^{-1}$ & Moment view, related to expectation coordinates by $M=\Sigma+\mu\mu^\top$. \\
\addlinespace
$\eta,\xi$ & $\eta=(h,-J/2)$, $\xi=(\mu,M)$ & Canonical natural and expectation charts. The separate symbol $\tau>0$ denotes a source-prior temperature, as in $\pi_{\tau,t}$. \\
\addlinespace
$\theta,\psi$ & Model and recognition parameters & Parameters of the normalized generative joint and normalized recognition law, respectively. \\
\addlinespace
$\ELBO,\VFE$ & Scalar functionals & Evidence lower bound and VFE 4.0 free energy, with $\VFE=-\ELBO$, under the stated support and integrability assumptions. \\
\end{longtable}
}
