\documentclass[12pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage[round]{natbib}
\usepackage{float}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tcolorbox}
\usepackage{enumitem}
\geometry{margin=1in}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\Sig}{\Sigma}
\newcommand{\SigQ}{\Sigma^q}
\newcommand{\SigP}{\Sigma^p}
\newcommand{\muQ}{\mu^q}
\newcommand{\muP}{\mu^p}
\newcommand{\Dmu}{\Delta\mu}
\newcommand{\vech}{\mathrm{vech}}
\newcommand{\Prec}{P}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{prediction}[theorem]{Prediction}

\title{The Inertia of Belief:\\Gauge-Covariant Variational Free Energy for Social Dynamics}

\author{
Robert C. Dennis\\
\texttt{cdenn016@gmail.com}
}

\date{\today}

\begin{document}
\maketitle



\begin{abstract}
We formulate an engineered gauge-covariant consensus energy for distribution-valued beliefs represented in heterogeneous local frames. Its relational weights follow from entropy-retaining attention optimization, whose rowwise elimination yields a reduced log-partition functional. Around local Gaussian consensus, the corresponding loss Hessian separates into prior, sensory, incoming relational, and outgoing relational or recoil stiffness. We take Fisher--Rao natural-gradient dynamics as the primary first-order update law. Belief inertia enters only through a separate kinetic-metric postulate: a positive local restriction of the stiffness may be adopted as kinetic geometry, but if the same functional supplies both kinetic mass $M$ and restoring Hessian $H_F$, then $M=H_F$ makes the generalized harmonic spectrum degenerate, $\omega^2=1$ up to scale. Under symmetric or reversible coupling, or at equilibrium under additional anchoring assumptions, the first-order theory has correspondences to restricted DeGroot and Friedkin--Johnsen dynamics and a finite-temperature analog of bounded confidence. In the stated unanchored, symmetric reciprocal two-cluster reduction, however, separated clusters retain nonzero mutual influence and are metastable rather than stably polarized. This construction is a local mathematical model: it distinguishes intrinsic Fisher geometry, loss curvature, and optional kinetic structure, and it treats social and psychological interpretations as mechanism-specific hypotheses rather than validated explanations.
\end{abstract}



\noindent\textbf{Keywords:} gauge-covariant variational free energy $\cdot$ opinion dynamics $\cdot$ information geometry $\cdot$ natural gradient $\cdot$ relational stiffness $\cdot$ belief inertia

\section{Introduction}

Social beliefs are often represented by point estimates in a common coordinate system. That representation suppresses two features that matter when heterogeneous agents exchange information: an agent may hold a distribution rather than a single opinion, and different agents may encode the same latent content in different local frames. Comparing such beliefs therefore requires both uncertainty-aware divergence and an explicit transport law. We study this problem for local Gaussian belief families, using gauge transport to compare probability distributions without identifying their coordinates by fiat.

The relevant landscape is already broad. Classical DeGroot and Friedkin--Johnsen models describe linear averaging and anchored influence \citep{degroot1974,friedkin1990}, while Hegselmann--Krause and Deffuant models impose bounded-confidence interaction rules \citep{hegselmann2002opinion,deffuant2000}. The closest direct opinion-inertia predecessor is Martins: his Bayesian-inspired opinion particles already possess force, inertia, and a harmonic-oscillator example \citep{Martins2015OpinionParticles}. Behavioral momentum instead operationalizes resistance to disruption in conditioned behavior and does not establish opinion-level Hamiltonian dynamics \citep{NevinMandellAtak1983}. Active-inference models supply further collective precedents. Federated inference treats communicated posteriors as evidence in shared-model inference \citep{friston2024federated}, whereas confirmation-biased policy selection can generate epistemic communities and echo chambers \citep{albarracin2022epistemic}; neither construction supplies the inter-frame transport studied here. Xue, Hirche, and Cao apply port-Hamiltonian systems and passivity to media-influenced opinion networks, but their state-space control model is not a gauge-coupled geometry of probabilistic beliefs \citep{XueHircheCao2020}.

Information-geometric mechanics also predates the present work. Pistone constructs Lagrangian dynamics on a finite-state statistical bundle \citep{Pistone2018StatisticalBundle}, and Chirco, Malag\`o, and Pistone develop Lagrangian and Hamiltonian probability dynamics that include accelerated natural-gradient motion \citep{ChircoMalagoPistone2022}. Leok and Zhang relate divergence functions to discrete Lagrangians under explicit geometric conditions rather than identifying every divergence with a unique kinetic energy \citep{LeokZhang2017}. Girolami and Calderhead use a position-dependent Fisher or Riemannian metric in Hamiltonian sampling, providing a clear precedent for metric-dependent kinetic geometry while keeping that metric conceptually distinct from objective curvature \citep{GirolamiCalderhead2011}. At the network level, periodically modulated second-order consensus already exhibits Laplacian-mode resonance \citep{BaumannSokolovTyloo2020}. Oscillatory social behavior is not diagnostic of inertia by itself: Sampson, Restrepo, and Porter obtain self-sustained and excitable opinion motion from feedback between individual and group opinions without the kinetic mechanism proposed here \citep{SampsonPorterRestrepo2025}.

Against these precedents, the gap is specific. We have not identified an earlier model that combines Gaussian KL consensus across heterogeneous local frames with an explicit local four-part stiffness decomposition into prior, sensory, incoming relational, and outgoing relational or recoil terms. The outgoing term is especially important because it records the curvature contributed when an agent serves as the source in other agents' relational energies. The contribution is this transported consensus and curvature construction, together with a strict separation between first-order inference and an optional kinetic extension; it is not the invention of opinion inertia, Hamiltonian probability dynamics, or oscillatory consensus.

The paper makes three contributions. First, it gives a gauge-covariant population consensus functional for distribution-valued beliefs in heterogeneous local frames, with a canonical log-partition reduction of attention. Second, it establishes a local Gaussian stiffness theorem separating prior, sensory, incoming relational, and outgoing relational/recoil curvature, including the reduced-Hessian covariance correction away from frozen consensus. Third, it gives a conditional kinetic extension and a proof-status map for social dynamics: symmetric/reversible DeGroot flow, restricted Friedkin--Johnsen equilibrium, soft bounded confidence, metastable clustering in the stated unanchored, symmetric reciprocal two-cluster reduction, and mechanism-specific falsification tests.

The scope is consequently narrow. This is a local Gaussian theory and model paper in mathematical sociophysics and computational social science, not a general psychology of belief perseverance or confirmation bias and not a participatory-physics theory. It does not infer adoption dynamics from individual belief trajectories; a diffusion claim would require an explicit adoption state and population hazard, as in the Bass model \citep{Bass1969}. Psychological constructs enter only as candidate mechanisms to be distinguished by controlled perturbations and matched alternative models.

\section{Epistemic Status and Scope}\label{sec:epistemic_status}

Table~\ref{tab:epistemic-status} fixes the claim-status contract for the manuscript. Each fence distinguishes a mathematical result or modeling choice from a stronger interpretation that the present construction does not establish.

\begin{table}[H]
\centering
\small
\caption{Epistemic-status fences governing the paper's claims.}
\label{tab:epistemic-status}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.08\textwidth}>{\raggedright\arraybackslash}p{0.84\textwidth}@{}}
\toprule
Fence & Claim status \\
\midrule
E1 & The population interaction term is an engineered consensus energy, not the evidence lower bound of a fixed population generative model. \\
E2 & The four-part stiffness is a local frozen-attention, consensus-point loss Hessian, not a global intrinsic metric. \\
E3 & The kinetic metric is an additional modeling postulate rather than a consequence of variational free energy. \\
E4 & Classical-model recovery is restricted to symmetric or reversible flows, or to equilibrium correspondences under stated assumptions. \\
E5 & In the stated unanchored, symmetric reciprocal two-cluster reduction, finite-temperature attractive attention supports metastable clustering, not exact stable polarization. \\
E6 & Psychological constructs are candidate mechanisms, not established explanations. \\
\bottomrule
\end{tabular}
\end{table}

Later definitions, propositions, and interpretations cite the applicable fence whenever their validity depends on one of these restrictions. The fence reference is part of the claim itself: dropping it does not strengthen the result, but changes it into an unsupported statement.

\section{Mathematical Framework}

\subsection{Gaussian Beliefs and Local Frames}

Agent $i$ carries a Gaussian belief $q_i=\mathcal N(\mu_i,\Sigma_i)$ and a quasi-static Gaussian prior $p_i=\mathcal N(\bar\mu_i,\bar\Sigma_i)$ on a common latent dimension $d$, with $\Sigma_i,\bar\Sigma_i\in\mathrm{SPD}(d)$. The coordinates are local: an invertible frame $U_i\in\mathrm{GL}(d)$ specifies how agent $i$ represents the latent variables. For any invertible affine map $T(x)=Ax+a$, Gaussian pushforward gives
\begin{equation}
T_{\#}\mathcal N(\mu,\Sigma)
=\mathcal N(A\mu+a,A\Sigma A^{\top}).
\label{eq:affine_gaussian_pushforward}
\end{equation}
The inter-agent transport used below is the linear specialization
\begin{equation}
\Omega_{ij}=U_iU_j^{-1},
\qquad
\Omega_{ij\#}q_j
=\mathcal N(\Omega_{ij}\mu_j,\Omega_{ij}\Sigma_j\Omega_{ij}^{\top}).
\label{eq:transport_cov}
\end{equation}
Thus both means and covariances are compared only after they have been expressed in the receiver's frame. This is the focused social-dynamics use of the gauge construction; no participatory-physics interpretation is required.

\begin{definition}[Transported edge energy]
The directed disagreement energy from source $j$ to receiver $i$ is
\begin{equation}
E_{ij}=\KL\!\left(q_i\,\middle\|\,\Omega_{ij\#}q_j\right),
\qquad
\Omega_{ij}=U_iU_j^{-1}.
\label{eq:transported_edge_energy}
\end{equation}
For Gaussians, this is the ordinary closed-form Gaussian KL evaluated using the transported mean and covariance in Eq.~\eqref{eq:transport_cov}.
\end{definition}

\begin{proposition}[Gauge covariance of the edge energy]
Under independent local frame changes $g_i\in\mathrm{GL}(d)$, let $q_i'=g_{i\#}q_i$ and $U_i'=g_iU_i$. Then $E_{ij}'=E_{ij}$.
\end{proposition}

\begin{proof}
The transformed transport is $\Omega_{ij}'=g_i\Omega_{ij}g_j^{-1}$, so functoriality of pushforward gives
\begin{equation}
\Omega_{ij\#}'q_j'
=(g_i\Omega_{ij}g_j^{-1})_{\#}(g_{j\#}q_j)
=g_{i\#}(\Omega_{ij\#}q_j).
\end{equation}
KL divergence is invariant when both arguments are pushed forward by the same invertible measurable map. Therefore
\begin{equation}
E_{ij}'
=\KL\!\left(g_{i\#}q_i\,\middle\|\,g_{i\#}(\Omega_{ij\#}q_j)\right)
=E_{ij}.
\end{equation}
This proves gauge covariance of the scalar edge energy. It does not imply that an ordinary coordinate Hessian is invariant under arbitrary nonlinear reparameterization; such Hessians are used later only as local affine-coordinate stiffnesses, consistently with fence E2 in Table~\ref{tab:epistemic-status}.
\end{proof}

\subsection{Canonical Entropy-Retaining Population Objective}

For each receiver $i$, let the attention-prior coefficients satisfy $\pi_{ij}>0$ and $\sum_j\pi_{ij}=1$ on its finite candidate-source set, let $\beta_i$ lie in the same probability simplex, and fix $\tau>0$. A hard mask is represented by removing an index from that set rather than assigning a positive-weight formula at $\pi_{ij}=0$. The authoritative scalar objective is
\begin{equation}
\boxed{
\mathcal F_{\mathrm{full}}
=\sum_i\KL(q_i\|p_i)
-\sum_i\mathbb E_{q_i}\log p(o_i\mid c_i)
+\sum_{i,j}\left[
\beta_{ij}E_{ij}
+\tau\beta_{ij}\log\frac{\beta_{ij}}{\pi_{ij}}
\right].}
\label{eq:canonical_full_vfe}
\end{equation}
The inter-agent block is an engineered, gauge-covariant consensus energy. In particular, $\mathcal F_{\mathrm{full}}$ is not claimed to be the negative evidence lower bound of one fixed population generative model over the original agent-state variables, as required by fence E1 in Table~\ref{tab:epistemic-status}. Its exact rowwise source-selection identity at unit categorical temperature is narrower than that population-level claim; Appendix~\ref{app:social_coupling} records the distinction.

\subsection{Optimized Attention and the Reduced Objective}

\begin{definition}[Rowwise source-independence assumption]
For the source-selection interpretation of a fixed row $i$, introduce a source label $z$ and factor the variational row as $Q_i(c,z=j)=q_i(c)\beta_{ij}$. Thus the focal belief $q_i(c)$ is independent of the source label under $Q_i$. During the row optimization, $q_i$, the transported source templates $\Omega_{ij\#}q_j$, the prior weights $\pi_{ij}$, and $\tau$ are held fixed. This is a substantive mean-field assumption at the mixture-to-softmax seam, not a statement that all agents' variational beliefs are fixed generative-model factors.
\end{definition}

The part of Eq.~\eqref{eq:canonical_full_vfe} that depends on row $i$ is
\begin{equation}
\Phi_i(\beta_i)
=\sum_j\left[
\beta_{ij}E_{ij}
+\tau\beta_{ij}\log\frac{\beta_{ij}}{\pi_{ij}}
\right],
\qquad
\sum_j\beta_{ij}=1.
\label{eq:attention_row_objective}
\end{equation}
Introducing a multiplier $\lambda_i$ for the simplex constraint gives the interior stationarity equations
\begin{equation}
E_{ij}+\tau\left(\log\frac{\beta_{ij}}{\pi_{ij}}+1\right)+\lambda_i=0.
\label{eq:attention_row_stationarity}
\end{equation}
Exponentiation and normalization therefore yield
\begin{equation}
\boxed{
\beta_{ij}^*=\frac{\pi_{ij}e^{-E_{ij}/\tau}}{Z_i},
\qquad
Z_i=\sum_k\pi_{ik}e^{-E_{ik}/\tau}.}
\label{eq:attention_weights}
\end{equation}
Strict convexity of the categorical KL term on the simplex makes this stationary point the unique minimizer. More directly, substituting Eq.~\eqref{eq:attention_weights} into Eq.~\eqref{eq:attention_row_objective} gives the exact decomposition
\begin{equation}
\Phi_i(\beta_i)
=-\tau\log Z_i+\tau\KL(\beta_i\|\beta_i^*).
\label{eq:attention_row_kl_decomposition}
\end{equation}
Consequently the optimized row and the fully reduced population objective are
\begin{align}
\mathcal F_{i,\mathrm{red}}&=-\tau\log Z_i,
\label{eq:canonical_row_reduced_vfe}\\
\mathcal F_{\mathrm{red}}
&=\sum_i\KL(q_i\|p_i)
-\sum_i\mathbb E_{q_i}\log p(o_i\mid c_i)
+\sum_i\mathcal F_{i,\mathrm{red}}.
\label{eq:canonical_reduced_vfe}
\end{align}

When the coefficients $\pi_{ij}$ and the temperature $\tau$ are fixed, direct differentiation of the log-partition function proves the envelope identity
\begin{equation}
\boxed{
d\mathcal F_{i,\mathrm{red}}
=-\tau\frac{dZ_i}{Z_i}
=\sum_j\beta_{ij}^*\,dE_{ij}.}
\label{eq:reduced_vfe_envelope}
\end{equation}
No derivative of $\beta_i^*$ appears in Eq.~\eqref{eq:reduced_vfe_envelope}; its contribution has already canceled through conditional minimization.

\subsection{Entropy-Suppressed Surrogate}

Dropping the categorical entropy and prior term after substituting optimized attention defines a different scalar,
\begin{equation}
S_i=\sum_j\beta_{ij}^*E_{ij}.
\label{eq:entropy_suppressed_surrogate}
\end{equation}
For fixed $\pi_{ij}$ and $\tau$, Eq.~\eqref{eq:attention_weights} gives
\begin{equation}
d\beta_{ij}^*
=-\tau^{-1}\beta_{ij}^*
\left(dE_{ij}-\mathbb E_{\beta_i^*}[dE_{ij}]\right).
\label{eq:attention_weight_differential}
\end{equation}
Applying the product rule to Eq.~\eqref{eq:entropy_suppressed_surrogate} then yields
\begin{equation}
\boxed{
dS_i
=\mathbb E_{\beta_i^*}[dE_{ij}]
-\tau^{-1}\operatorname{Cov}_{\beta_i^*}(E_{ij},dE_{ij}).}
\label{eq:entropy_suppressed_differential}
\end{equation}
Here the covariance between the scalar $E_{ij}$ and the one-form $dE_{ij}$ is understood componentwise. It has no universal sign in a direction identified with increasing or decreasing social similarity. The surrogate response therefore neither equals the canonical envelope response in general nor implies a sign-definite strengthening of similarity selection.

\subsection{Adaptive Prior Precision}
\label{sec:state_dependent_precision}

The canonical objective in Eq.~\eqref{eq:canonical_full_vfe} uses unit prior weight. Its adaptive-precision extension replaces the $i$th prior summand by the complete sector
\begin{equation}
\Psi_i(\alpha_i,D_i)
=\alpha_iD_i+R(\alpha_i),
\qquad
D_i=\KL(q_i\|p_i),
\qquad
R(\alpha_i)=b_0\alpha_i-c_0\log\alpha_i,
\label{eq:adaptive_self_coupling}
\end{equation}
where $b_0,c_0>0$ and $\alpha_i>0$. Because
\begin{equation}
\partial_{\alpha_i}\Psi_i
=D_i+b_0-\frac{c_0}{\alpha_i},
\qquad
\partial_{\alpha_i}^2\Psi_i=\frac{c_0}{\alpha_i^2}>0,
\end{equation}
the unique conditional optimum is
\begin{equation}
\boxed{\alpha_i^*=\frac{c_0}{b_0+D_i}.}
\label{eq:state_dependent_alpha}
\end{equation}

Two derivatives that have previously been conflated belong to different scalar objectives. For the bare product $B_i(D)=\alpha_i^*(D)D$ and the reduced regularized sector $\bar\Psi_i(D)=\Psi_i(\alpha_i^*(D),D)$, respectively,
\begin{align}
\partial_D(\alpha_i^*D)
&=\frac{b_0c_0}{(b_0+D)^2},
\label{eq:adaptive_bare_product_derivative}\\
\partial_D[\alpha_i^*D+R(\alpha_i^*)]
&=\frac{c_0}{b_0+D}=\alpha_i^*.
\label{eq:adaptive_envelope_derivative}
\end{align}
The second identity is the envelope theorem. Except at $D=0$, Eqs.~\eqref{eq:adaptive_bare_product_derivative} and~\eqref{eq:adaptive_envelope_derivative} produce different force coefficients and therefore generate distinct dynamics.

Let $\bar\Lambda_{p_i}=\bar\Sigma_i^{-1}$ and $\Lambda_{q_i}=\Sigma_i^{-1}$. At fixed $\alpha_i$, the adaptive sector contributes
\begin{align}
\nabla_{\mu_i}\Psi_i
&=\alpha_i\bar\Lambda_{p_i}(\mu_i-\bar\mu_i),
\label{eq:adaptive_prior_mean_gradient}\\
\nabla_{\Sigma_i}\Psi_i
&=\frac{\alpha_i}{2}(\bar\Lambda_{p_i}-\Lambda_{q_i}).
\label{eq:adaptive_prior_covariance_gradient}
\end{align}
For the reduced regularized sector, the same formulas hold with $\alpha_i=\alpha_i^*$ by Eq.~\eqref{eq:adaptive_envelope_derivative}. Thus the prior coefficient is carried consistently through both the mean and covariance gradients; any corresponding fixed-$\alpha_i$ prior Hessian block carries the same factor.

The response of $\alpha_i^*$ matters when one differentiates the reduced regularized sector a second time. Holding $\Sigma_i$, $p_i$, $b_0$, and $c_0$ fixed, and writing $g_i=\bar\Lambda_{p_i}(\mu_i-\bar\mu_i)$, its mean-coordinate Hessian is
\begin{equation}
\nabla_{\mu_i}^2\bar\Psi_i
=\alpha_i^*\bar\Lambda_{p_i}
-\frac{(\alpha_i^*)^2}{c_0}g_ig_i^{\top}.
\label{eq:adaptive_prior_reduced_hessian}
\end{equation}
The negative rank-one term in Eq.~\eqref{eq:adaptive_prior_reduced_hessian} belongs specifically to the mean block of the envelope-reduced, regularized prior sector. It is not a correction to the bare product, the covariance block, the full population Hessian, or an intrinsic Fisher metric. Its possible loss of positive definiteness therefore limits only an interpretation that treats this reduced coordinate curvature itself as a positive stiffness; it does not invalidate the scalar objective.

%==============================================================================
\subsection{Fisher Arc Length as an Information-Theoretic Clock}
\label{sec:proper_time}
%==============================================================================

A fundamental issue in any dynamical theory of belief is the definition of time. Wall-clock time is unsuitable because cognitive processes operate on different timescales for different agents and contexts. We propose defining a Fisher arc-length clock as the information-theoretic arc length traversed on the statistical manifold whereby time is taken as "a difference which makes a difference". We use the symbol $\ell$ for this clock to keep it distinct from the attention temperature $\tau$ and from the dynamical relaxation times introduced later.

For an infinitesimal belief change $d\mu$, the arc-length increment is

\begin{equation}
\boxed{d\ell = \sqrt{d\mu^T \Sigma^{-1} d\mu} = \|d\mu\|_{\Sigma^{-1}}}
\end{equation}

This definition has several appealing properties:

\paragraph{Scale dependence.} For a high-precision agent ($\Sigma$ small, $\Sigma^{-1}$ large), a small change in $\mu$ corresponds to a large arc-length increment such that "time moves fast" in the sense that each update is cognitively significant. For a low-precision agent, the same parametric change corresponds to a small arc-length increment whereby updates are relatively insignificant or slow. A single bit of information is enormous for a simple, small-scale agent but imperceptible for a complex one such as humans.

\paragraph{Information-theoretic interpretation.} To second order, both KL directions give the same result

\begin{equation}
\text{KL}(q + dq \| q) \approx \text{KL}(q \| q + dq) \approx \frac{1}{2} d\mu^T \Sigma^{-1} d\mu = \frac{1}{2} d\ell^2 .
\end{equation}

Thus the arc-length clock measures accumulated information change. The choice between KL directions is immaterial at this order.

\paragraph{Invariance.} The Fisher arc length is invariant under reparameterization of the belief space, depending only on the intrinsic geometry of the statistical manifold.

For a trajectory $\mu(t)$ parameterized by some external parameter $t$, the total arc length elapsed is

\begin{equation}
\ell = \int \sqrt{\dot{\mu}^T \Sigma^{-1} \dot{\mu}}\ dt .
\end{equation}

This Fisher arc-length clock plays a role loosely comparable to proper time in special relativity, in that different agents accumulate it at different rates depending on their precision; we draw the comparison only as an analogy and do not claim a Lorentzian metric structure, which would require a derived signature and is not established here. The velocity through belief space determines how quickly arc length accumulates, with high-precision agents accumulating it faster for identical parametric motion. The intrinsic metric that defines this arc length does not, by itself, select a kinetic energy or an inertial mass.

%==============================================================================
\subsection{Fisher--Rao Natural Gradient as the Primary Dynamics}
\label{sec:natural_gradient_dynamics}
%==============================================================================

The primary belief dynamics are first order.  For each Gaussian factor, the inverse Fisher metric in affine $(\mu_i,\Sigma_i)$ coordinates maps the Euclidean coordinate derivatives of the canonical objective to
\begin{equation}
\boxed{
\dot\mu_i=-\eta_\mu\Sigma_i\nabla_{\mu_i}\mathcal F,
\qquad
\dot\Sigma_i=-2\eta_\Sigma\Sigma_i
(\nabla_{\Sigma_i}\mathcal F)\Sigma_i,}
\label{eq:primary_fisher_flow}
\end{equation}
with $\eta_\mu,\eta_\Sigma>0$ and symmetric covariance derivative.  Equation~\eqref{eq:primary_fisher_flow} is the Fisher--Rao steepest-descent flow on the Gaussian statistical manifold, not inverse-loss-Hessian or Newton flow \citep{Amari2016}.  In particular, the four-part loss stiffness derived below does not replace the intrinsic Fisher preconditioner in Eq.~\eqref{eq:primary_fisher_flow}.

An overdamped mechanical analogy reproduces this flow only after a tensor friction has been specified.  If one separately postulates a kinetic block $\mathbf M=\mathbf G$ and writes $\mathbf M\ddot x+\boldsymbol\Gamma\dot x+\nabla_x\mathcal F=0$, then choosing $\boldsymbol\Gamma=\mathbf M/\eta$ gives $\dot x=-\eta\mathbf G^{-1}\nabla_x\mathcal F$ on the slow manifold.  Distinct mean and covariance learning rates give the corresponding blockwise choices $\boldsymbol\Gamma^\mu=\mathbf G^\mu/\eta_\mu$ and $\boldsymbol\Gamma^\Sigma=\mathbf G^\Sigma/\eta_\Sigma$.  For a general independently chosen kinetic metric $\mathbf M\ne\mathbf G$, the choice $\boldsymbol\Gamma=\mathbf M/\eta$ instead produces an $\mathbf M$-metric flow and must not be called the Fisher natural gradient.  A scalar damping coefficient cannot establish either identification.


%------------------------------------------------------------------------------
\subsection{Local Gaussian Stiffness and the Three Geometries}
%------------------------------------------------------------------------------

Let $x=(\mu_1,\ldots,\mu_N,\Sigma_1,\ldots,\Sigma_N)$ denote affine Gaussian coordinates. Three geometric objects must be distinguished. The intrinsic Fisher--Rao metric $\mathbf G(q)$ depends only on the statistical family and the point $q$. The coordinate loss Hessian
\begin{equation}
\mathbf H_F(x)=\nabla_x^2\mathcal F(x)
\label{eq:loss_hessian_definition}
\end{equation}
is the local stiffness of the selected scalar objective. Outside a critical point it is an affine-coordinate stiffness, not a tensor under arbitrary nonlinear reparameterizations. The symbol $\mathbf M(x)$ is reserved for a positive kinetic metric introduced only if the kinetic postulate is adopted. Thus $\mathbf G$, $\mathbf H_F$, and $\mathbf M$ are not interchangeable notation.

The {\v C}encov--Chentsov theorem justifies $\mathbf G$, up to scale, as the classical monotone metric under sufficient-statistic maps \citep{Cencov1982}; it does not identify the Hessian of a prior, likelihood, and relational objective with that intrinsic metric. A matched KL term has the Fisher quadratic form at its minimum, and a fixed-covariance Gaussian KL is quadratic in its mean. Consequently $\mathbf H_F=\mathbf G$ can hold on an explicitly identified matching block or in a declared quadratic regime, but it does not hold globally for the summed population objective merely because both objects contain precision matrices.

\begin{theorem}[Four-part local Gaussian stiffness]
\label{thm:four_part_stiffness}
Consider the objective in Eq.~\eqref{eq:canonical_full_vfe}, with the Gaussian observation model used in this paper. Fix the transports and covariances. Evaluate the optimized attention weights in Eq.~\eqref{eq:attention_weights}, then freeze those values while differentiating the Gaussian coordinates. Also hold $\alpha_i$ fixed, with $\alpha_i=1$ in the canonical prior sector and $\alpha_i=\alpha_i^*$ when a conditionally optimized adaptive value is frozen. At a gauge-consensus point $q_i=\Omega_{ik\#}q_k$ on every active edge, the mean-sector block of the full stacked loss Hessian has diagonal blocks
\begin{equation}
\boxed{
[\mathbf H_{\mu\mu}]_{ii}
=\underbrace{\alpha_i\bar\Lambda_{p_i}}_{\text{prior stiffness}}
+\underbrace{\Lambda_{o_i}}_{\text{sensory stiffness}}
+\underbrace{\sum_k\beta_{ik}\widetilde\Lambda_{q_k}}_{\text{incoming relational stiffness}}
+\underbrace{\sum_j\beta_{ji}\Lambda_{q_i}}_{\text{outgoing relational/recoil stiffness}},}
\label{eq:stiffness_diagonal}
\end{equation}
where $\widetilde\Lambda_{q_k}=\Omega_{ik}^{-\top}\Lambda_{q_k}\Omega_{ik}^{-1}$. For $i\ne k$, its off-diagonal sender--receiver blocks are
\begin{equation}
\boxed{
[\mathbf H_{\mu\mu}]_{ik}
=-\beta_{ik}\Omega_{ik}^{-\top}\Lambda_{q_k}
-\beta_{ki}\Lambda_{q_i}\Omega_{ki}^{-1}.}
\label{eq:stiffness_offdiagonal}
\end{equation}
The assembled matrix is symmetric, including when the frozen attention matrix is asymmetric.
\end{theorem}

\begin{proof}
For one directed edge from sender $k$ to receiver $i$, write $r_{ik}=\mu_i-\Omega_{ik}\mu_k$. The mean-dependent part of its Gaussian KL is
\begin{equation}
E_{ik}^{(\mu)}=\frac12 r_{ik}^{\top}\widetilde\Lambda_{q_k}r_{ik}.
\label{eq:edge_mean_quadratic}
\end{equation}
Direct differentiation gives
\begin{align}
\nabla_{\mu_i}E_{ik}^{(\mu)}
&=\widetilde\Lambda_{q_k}r_{ik},
&
\nabla_{\mu_k}E_{ik}^{(\mu)}
&=-\Omega_{ik}^{\top}\widetilde\Lambda_{q_k}r_{ik},
\label{eq:edge_mean_gradients}\\
\nabla_{\mu_i\mu_i}^2E_{ik}^{(\mu)}
&=\widetilde\Lambda_{q_k},
&
\nabla_{\mu_k\mu_k}^2E_{ik}^{(\mu)}
&=\Lambda_{q_k},
\label{eq:edge_mean_diagonal_hessians}\\
\nabla_{\mu_i\mu_k}^2E_{ik}^{(\mu)}
&=-\Omega_{ik}^{-\top}\Lambda_{q_k},
&
\nabla_{\mu_k\mu_i}^2E_{ik}^{(\mu)}
&=-\Lambda_{q_k}\Omega_{ik}^{-1}.
\label{eq:edge_mean_cross_hessians}
\end{align}
The prior and observation quadratics contribute $\alpha_i\bar\Lambda_{p_i}$ and $\Lambda_{o_i}$ to the $i$th diagonal block. Every edge received by $i$ contributes its transported sender precision, while every edge on which $i$ is the sender contributes its untransported precision $\Lambda_{q_i}$. Summing these terms proves Eq.~\eqref{eq:stiffness_diagonal}. For $i\ne k$, the edge received by $i$ contributes the first term of Eq.~\eqref{eq:stiffness_offdiagonal}, and the reciprocal edge received by $k$ contributes the second. Equation~\eqref{eq:edge_mean_cross_hessians} then gives $[\mathbf H_{\mu\mu}]_{ki}=[\mathbf H_{\mu\mu}]_{ik}^{\top}$ without requiring $\beta_{ik}=\beta_{ki}$.

The same calculation exposes the full frozen-attention force balance:
\begin{align}
\nabla_{\mu_i}\mathcal F_{\mathrm{full}}
={}&\alpha_i\bar\Lambda_{p_i}(\mu_i-\bar\mu_i)
+\Lambda_{o_i}(\mu_i-o_i)\nonumber\\
&+\sum_k\beta_{ik}\widetilde\Lambda_{q_k}
(\mu_i-\Omega_{ik}\mu_k)
+\sum_j\beta_{ji}\Lambda_{q_i}
(\mu_i-\Omega_{ji}^{-1}\mu_j).
\label{eq:stiffness_force_balance}
\end{align}
Differentiating this stacked force reproduces both block formulas. At consensus the mean--covariance cross blocks of the edge KLs vanish, so the displayed mean sector decouples locally from covariance variations. This completes the proof.
\end{proof}

The theorem concerns the complete scalar and includes both receiver and sender derivatives. The outgoing term is therefore a recoil contribution to local curvature, not accumulated memory and not an inertial effect by itself. If the envelope-reduced adaptive prior of Eq.~\eqref{eq:adaptive_prior_reduced_hessian} is differentiated rather than frozen, its stated negative rank-one response must be added to the prior mean block.

\begin{proposition}[Reduced-attention Hessian correction]
\label{prop:reduced_attention_hessian}
For any affine coordinate vector on which the row energies depend, the exact Hessian of the optimized row in Eq.~\eqref{eq:canonical_row_reduced_vfe} is
\begin{equation}
\boxed{
\nabla^2\mathcal F_{i,\mathrm{red}}
=\mathbb E_{\beta_i^*}[\nabla^2E_{ij}]
-\tau^{-1}\operatorname{Cov}_{\beta_i^*}
(\nabla E_{ij},\nabla E_{ij}).}
\label{eq:reduced_row_hessian}
\end{equation}
Here the covariance is the matrix-valued covariance of the edge gradients.
\end{proposition}

\begin{proof}
Differentiate the envelope identity in Eq.~\eqref{eq:reduced_vfe_envelope} and substitute Eq.~\eqref{eq:attention_weight_differential}. The derivative of each edge gradient gives the expected edge Hessian, while the derivative of the Gibbs weights gives the negative covariance term in Eq.~\eqref{eq:reduced_row_hessian}.
\end{proof}

The covariance term is positive semidefinite before its minus sign, so attention response can reduce curvature and can create indefinite directions away from consensus. No global positive-definiteness claim follows for the reduced Hessian. At exact edgewise consensus the KL gradients vanish and so does this covariance correction; freezing the optimized weights omits the correction by construction.

\begin{proposition}[Locally positive covariance diagonal block]
\label{prop:covariance_stiffness}
Under the same frozen-attention and edgewise-consensus assumptions as Theorem~\ref{thm:four_part_stiffness}, the clean diagonal covariance block, represented on symmetric matrix coordinates, is
\begin{equation}
\boxed{
[\mathbf H_{\Sigma\Sigma}]_{ii}
=\frac12(\Lambda_{q_i}\otimes\Lambda_{q_i})
\left(\alpha_i+\sum_k\beta_{ik}+\sum_j\beta_{ji}\right).}
\label{eq:stiffness_covariance}
\end{equation}
The sensory term contributes no covariance Hessian because it is affine in $\Sigma_i$.
\end{proposition}

\begin{proof}
For a symmetric covariance variation $U$, each matched Gaussian KL in which $q_i$ is receiver or sender has second variation $\frac12\operatorname{tr}(\Lambda_{q_i}U\Lambda_{q_i}U)$. The prior supplies the same form with coefficient $\alpha_i$, incoming edges supply coefficients $\beta_{ik}$, and outgoing edges supply coefficients $\beta_{ji}$. Their sum gives Eq.~\eqref{eq:stiffness_covariance}. Since
\begin{equation}
\operatorname{tr}(\Lambda_{q_i}U\Lambda_{q_i}U)
=\|\Lambda_{q_i}^{1/2}U\Lambda_{q_i}^{1/2}\|_F^2,
\end{equation}
this diagonal block is positive definite on nonzero symmetric $U$ whenever its scalar coefficient is positive. This is a local block statement only: it does not establish positive definiteness of the full stacked covariance Hessian or of the envelope-reduced Hessian.
\end{proof}

%------------------------------------------------------------------------------
\subsection{Kinetic Geometry as a Conditional Extension}
\label{sec:hamiltonian}
%------------------------------------------------------------------------------

The objective determines forces and local loss curvature, whereas a second-order theory additionally requires a positive kinetic metric.  We reserve $\mathbf M(x)$ for that metric.  Statistical-bundle Lagrangians and Hamiltonian probability dynamics provide precedents for such an additional construction, including accelerated natural-gradient mechanics, but they begin by specifying the mechanical structure rather than deriving a unique kinetic energy from a divergence \citep{Pistone2018StatisticalBundle,ChircoMalagoPistone2022}.  Likewise, the information-geometric/discrete-mechanical bridge of Leok and Zhang holds under explicit Hessian-manifold and variational conditions, and Riemann-manifold Hamiltonian Monte Carlo chooses a position-dependent Fisher or Riemannian kinetic geometry for an algorithmic purpose \citep{LeokZhang2017,GirolamiCalderhead2011}.  These precedents motivate, but do not prove, the following assumption.

\begin{tcolorbox}[colback=gray!5,colframe=gray!75,title=Kinetic-metric postulate]
\textbf{Assumption.} In a reciprocal, frozen-attention, local-consensus regime, a positive restriction of the gauge-VFE stiffness may be adopted as a kinetic metric for belief-configuration change.
\end{tcolorbox}

This is a modeling choice, not a theorem of variational free energy or information geometry.  It applies only after the chosen restriction is positive definite; it does not identify the intrinsic Fisher metric $\mathbf G$, the affine-coordinate loss Hessian $\mathbf H_F$, and the kinetic metric $\mathbf M$ globally.

The postulate also exposes a degeneracy that any oscillator interpretation must confront.  If the same local objective supplies the restoring stiffness, a normal mode $v\ne0$ solves
\begin{equation}
\mathbf H_Fv=\omega^2\mathbf Mv.
\label{eq:hessian_mass_spectrum}
\end{equation}
Consequently,
\begin{equation}
\boxed{
\mathbf M=\mathbf H_F
\quad\Longrightarrow\quad
\omega^2=1,}
\label{eq:hessian_mass_degeneracy}
\end{equation}
up to the declared conversion scale (more generally, $\mathbf M=c\mathbf H_F$ gives $\omega^2=c^{-1}$).  Thus the four-part stiffness spectrum cannot simultaneously serve as an identical mass spectrum and yield nontrivial mode-dependent frequencies.  Such dependence requires an operationally independent kinetic metric and restoring stiffness, for example a frozen pre-intervention $\mathbf M$ followed by manipulated evidence curvature, or the intrinsic Fisher metric $\mathbf G$ paired with a distinct local objective Hessian $\mathbf H_F$.  The conditional formulas below therefore use a positive kinetic metric $\mathbf M$ and a separately specified positive restoring stiffness $\mathbf K$.

\section{Results}

%==============================================================================
\subsection{Conditional Kinetic Mechanics and Full Coupled Momentum}
\label{sec:cognitive-momentum}
%==============================================================================

The second-order extension is conditional on the postulate in Section~\ref{sec:hamiltonian}; it is not the primary inference law.  Opinion particles with force and inertia, including a harmonic example, already appear in Martins's Bayesian-inspired construction, while Xue, Hirche, and Cao use port-Hamiltonian and passivity methods for a controlled opinion network \citep{Martins2015OpinionParticles,XueHircheCao2020}.  The present mechanism is narrower: it applies a separately declared kinetic metric to the gauge-transported VFE potential.

Let $\boldsymbol\mu=(\mu_1^\top,\ldots,\mu_N^\top)^\top$ and let the locally constant mean-sector kinetic metric have blocks $\mathbf M_{ik}^\mu$.  The Lagrangian and conjugate momentum are
\begin{equation}
L=\frac12\dot{\boldsymbol\mu}^{\top}\mathbf M^\mu
\dot{\boldsymbol\mu}-\mathcal F(\boldsymbol\mu),
\qquad
\boldsymbol\pi^\mu
=\frac{\partial L}{\partial\dot{\boldsymbol\mu}}
=\mathbf M^\mu\dot{\boldsymbol\mu}.
\label{eq:conditional_mean_lagrangian}
\end{equation}
Therefore the momentum assigned to agent $i$ is the full coupled row
\begin{equation}
\boxed{
\pi_i^\mu=\sum_k\mathbf M_{ik}^\mu\dot\mu_k.}
\label{eq:full_coupled_momentum}
\end{equation}
The simpler expression $\pi_i^\mu=M_i\dot\mu_i$ is valid only under an explicit block-diagonal approximation $\mathbf M_{ik}^\mu=\delta_{ik}M_i$.  In particular, the four diagonal stiffness contributions in Eq.~\eqref{eq:stiffness_diagonal} do not separately define four momenta, and off-diagonal sender--receiver blocks cannot be dropped when the full coupled kinetic metric is used.  Covariance motion has the analogous cotangent variable $\Pi_i^\Sigma$ described in Appendix~\ref{app:hamiltonian}.

For positive $\mathbf M$ the corresponding local Hamiltonian is
\begin{equation}
\boxed{
H(\boldsymbol\mu,\boldsymbol\pi)
=\frac12\boldsymbol\pi^{\top}\mathbf M^{-1}\boldsymbol\pi
+\mathcal F(\boldsymbol\mu).}
\label{eq:conditional_local_hamiltonian}
\end{equation}
If $\mathbf M$ varies with the state, Hamilton's momentum equation includes the metric-derivative term given in Appendix~\ref{app:hamiltonian}; the constant-metric oscillator laws below do not apply without that term.

\subsection{Belief Overshoot and Perseverance}\label{sec:overshoot}

Linearize about a local equilibrium and independently specify a positive kinetic metric $\mathbf M$, a positive restoring stiffness $\mathbf K$, and a positive-semidefinite damping tensor $\boldsymbol\Gamma$.  The conditional driven equation is
\begin{equation}
\mathbf M\ddot{\delta\boldsymbol\mu}
+\boldsymbol\Gamma\dot{\delta\boldsymbol\mu}
+\mathbf K\delta\boldsymbol\mu
=f(t).
\label{eq:conditional_matrix_oscillator}
\end{equation}
For a nonzero modal vector $v$, set $\delta\boldsymbol\mu=zv$ and define
\begin{equation}
m=v^\top\mathbf Mv,
\qquad
k=v^\top\mathbf Kv,
\qquad
\gamma=v^\top\boldsymbol\Gamma v,
\qquad
f_v(t)=v^\top f(t).
\label{eq:conditional_modal_scalars}
\end{equation}
Premultiplying Eq.~\eqref{eq:conditional_matrix_oscillator} by $v^\top$ yields
\begin{equation}
\boxed{m\ddot z+\gamma\dot z+kz=f_v(t).}
\label{eq:conditional_modal_oscillator}
\end{equation}
If $v$ satisfies $\mathbf Kv=\omega_0^2\mathbf Mv$, then $k=\omega_0^2m$.  An exact one-mode reduction additionally requires the span of $v$ to be invariant under the relevant operators; otherwise this is a projected modal equation.  These hypotheses are essential: a state-dependent metric, evolving attention, indefinite reduced Hessian, or nonmodal damping requires the full coupled equations rather than the scalar formulas below.

In the nearly conservative harmonic case, set $f_v=0$ during the coast and neglect damping over the stopping interval.  Energy conversion gives
\begin{equation}
\frac12m v_0^2=\frac12k d_{\mathrm{over}}^2,
\qquad
\boxed{d_{\mathrm{over}}=|v_0|\sqrt{\frac{m}{k}}.}
\label{eq:conditional_harmonic_overshoot}
\end{equation}
Thus $d_{\mathrm{over}}\propto\sqrt m$ only when the initial velocity $v_0$, restoring stiffness $k$, and damping protocol are held fixed.  If initial momentum $p_0=mv_0$ is held fixed instead, then $d_{\mathrm{over}}=|p_0|/\sqrt{mk}$ decreases as $m^{-1/2}$.  Neither scaling follows from the four-part loss stiffness unless the kinetic metric and the post-intervention restoring stiffness have been independently operationalized.

A different protocol applies a constant opposing force of magnitude $f_0>0$ with no spring.  Integrating $m\ddot z=-f_0$ until the velocity reaches zero yields
\begin{equation}
\boxed{
d_{\mathrm{stop}}=\frac{m v_0^2}{2f_0}
=\frac{p_0^2}{2mf_0}.}
\label{eq:conditional_ballistic_stop}
\end{equation}
The first form is linear in $m$ at fixed initial velocity and force; the second is inverse in $m$ at fixed initial momentum and force.  The harmonic overshoot and constant-force stopping laws are distinct interventions and must not be used interchangeably.

%------------------------------------------------------------------------------
\subsection{Belief Oscillation}
%------------------------------------------------------------------------------

Equation~\eqref{eq:conditional_matrix_oscillator}, rather than the primary Fisher flow, supplies the conditional oscillator.  Second-order opinion motion is not novel by itself: Martins gives an inferential opinion-particle harmonic oscillator, and network-level second-order consensus can exhibit Laplacian-mode parametric resonance under periodic coupling \citep{Martins2015OpinionParticles,BaumannSokolovTyloo2020}.  The distinctive question here is whether an independently defined kinetic metric improves predictions once the gauge-VFE restoring stiffness is manipulated.

\subsubsection{The Damped Epistemic Oscillator}

For one modal coordinate, the characteristic roots of $m\ddot z+\gamma\dot z+kz=0$ are
\begin{equation}
r_\pm=\frac{-\gamma\pm\sqrt{\gamma^2-4mk}}{2m}.
\label{eq:conditional_modal_roots}
\end{equation}
When $\gamma^2<4mk$, the damped angular frequency and envelope time are
\begin{equation}
\boxed{
\omega_d=\sqrt{\frac{k}{m}-\frac{\gamma^2}{4m^2}},
\qquad
\tau_{\mathrm{env}}=\frac{2m}{\gamma}.}
\label{eq:conditional_damped_frequency}
\end{equation}
These dependences compare modes or interventions at fixed $k$ and $\gamma$.  Holding a damping ratio rather than $\gamma$ fixed changes the scaling.  Moreover, setting $m=k$ because both are the same loss Hessian collapses the undamped frequency to the unit value already shown in Eq.~\eqref{eq:hessian_mass_degeneracy}.

\subsubsection{Three Dynamical Regimes}

The discriminant $\Delta=\gamma^2-4mk$ yields overdamped ($\Delta>0$), critically damped ($\Delta=0$), and underdamped ($\Delta<0$) conditional regimes.  These are exact statements about the postulated scalar mechanical mode.  They are neither consequences of information geometry alone nor evidence that observed nonmonotonic opinion trajectories uniquely identify this mechanism; feedback systems and time-dependent network coupling can also oscillate or resonate.

\subsection{Resonant Response to Periodic Evidence}

For a sinusoidal modal force $f_v(t)=f_0\cos(\omega t)$, with force amplitude $f_0$, mass $m$, stiffness $k$, and damping $\gamma$ all held fixed while the driving frequency is varied, the steady-state displacement amplitude is
\begin{equation}
A(\omega)=\frac{f_0}
{\sqrt{(k-m\omega^2)^2+(\gamma\omega)^2}}.
\label{eq:conditional_resonance_amplitude}
\end{equation}
Differentiating $A^{-2}$ shows that a positive-frequency displacement peak exists when $\gamma^2<2mk$ and occurs at
\begin{equation}
\boxed{
\omega_{\mathrm{res}}
=\sqrt{\frac{k}{m}-\frac{\gamma^2}{2m^2}}.}
\label{eq:conditional_resonance_frequency}
\end{equation}
This is a forced-response result under a fixed-force protocol.  The velocity-amplitude response instead peaks at $\sqrt{k/m}$, and a periodically modulated coupling produces the parametric network resonance studied by Baumann, Sokolov, and Tyloo rather than the additive-force response in Eq.~\eqref{eq:conditional_resonance_amplitude} \citep{BaumannSokolovTyloo2020}.


\subsubsection{Amplitude at Resonance}

Substitution of Eq.~\eqref{eq:conditional_resonance_frequency} gives the exact peak
\begin{equation}
\boxed{
A_{\mathrm{res}}
=\frac{f_0}{\gamma\sqrt{k/m-\gamma^2/(4m^2)}}.}
\label{eq:conditional_resonance_peak}
\end{equation}
Only in the light-damping limit does this reduce to $(f_0/\gamma)\sqrt{m/k}$.  That square-root comparison holds at fixed force amplitude, stiffness, and damping coefficient; it does not hold at fixed acceleration amplitude $f_0/m$, fixed input energy, or fixed damping ratio.  Because $m$ and $k$ must be operationally independent, the formula cannot be interpreted by substituting the same four-part loss Hessian for both.


%------------------------------------------------------------------------------
\subsection{Belief Perseverance in Social Embeddings}
%------------------------------------------------------------------------------

When the restoring force and external drive are removed after an intervention, the conditional modal equation becomes $m\ddot z+\gamma\dot z=0$.  Its velocity decays as
\begin{equation}
\dot z(t)=\dot z(0)e^{-t/\tau_{\mathrm{coast}}},
\qquad
\boxed{\tau_{\mathrm{coast}}=\frac{m}{\gamma}.}
\label{eq:conditional_coasting_time}
\end{equation}
Thus the coasting time is proportional to $m$ only when the damping coefficient $\gamma$ is held fixed.  Under a fixed damping ratio $\zeta=\gamma/(2\sqrt{mk})$, it scales instead as $\sqrt{m/k}$.  The coasting time in Eq.~\eqref{eq:conditional_coasting_time} is also distinct from the underdamped displacement-envelope time $2m/\gamma$ in Eq.~\eqref{eq:conditional_damped_frequency} and from the spring-dominated overdamped slow time $\gamma/k$.  None of these times may be inferred directly from precision without an independently specified kinetic metric and damping protocol.

\subsubsection{The Debunking Problem}

Interpreting this mathematical coast as belief perseverance is a falsifiable mechanism-specific hypothesis, not a consequence of the primary Fisher flow.  A test would have to manipulate or estimate $m$ separately from $k$ and $\gamma$ and then compare the observed post-drive velocity decay with Eq.~\eqref{eq:conditional_coasting_time}.


%------------------------------------------------------------------------------
\subsection{Sociology and Multi-Agent Momentum Transfer}
%------------------------------------------------------------------------------

The complete coupled momentum balance follows from the scalar potential and the full momentum in Eq.~\eqref{eq:full_coupled_momentum}.  This differs from a receiver-only influence update: every directed edge contributes both a receiver force and a sender recoil in the sender's cotangent frame.

\subsubsection{Coupled Equations of Motion}

For locally constant $\mathbf M$ and a damping tensor $\boldsymbol\Gamma$, the stacked equation is
\begin{equation}
\boxed{
\dot{\boldsymbol\pi}
=-\nabla_{\boldsymbol\mu}\mathcal F
-\boldsymbol\Gamma\dot{\boldsymbol\mu}
+f_{\mathrm{ext}},
\qquad
\boldsymbol\pi=\mathbf M\dot{\boldsymbol\mu}.}
\label{eq:full_coupled_momentum_balance}
\end{equation}
This formula remains coupled even if damping is block diagonal.  For state-dependent $\mathbf M$, the Hamiltonian metric-derivative force must be restored, so Eq.~\eqref{eq:full_coupled_momentum_balance} is the local constant-metric form rather than a global identity.

\paragraph{Dynamic attention.} The expansion above assumes fixed attention weights. When attention is optimized inside the entropy-retaining row objective, its canonical gradient is the envelope response in Eq.~\eqref{eq:reduced_vfe_envelope}, with no additional first-order attention-reallocation term. Applying a product rule after dropping the categorical term instead differentiates the distinct surrogate $S_i$ and produces Eq.~\eqref{eq:entropy_suppressed_differential}. The covariance correction in that equation has no universal homophily sign and cannot by itself establish accelerated polarization.

\paragraph{Regime of validity.} When attention weights or the postulated kinetic metric vary with the belief state, the Hamiltonian is generally nonseparable.  The modal formulas above apply only after $\mathbf M$, $\mathbf K$, and $\boldsymbol\Gamma$ have been frozen locally and the resulting positive modes identified.  Large-amplitude or rapidly adapting trajectories require the full state-dependent equations.


\subsubsection{Epistemic Momentum Transfer}


For the directed edge $k\to i$, define the receiver-frame current
\begin{equation}
\boxed{
J_{k\to i}
=\beta_{ik}\widetilde\Lambda_{q_k}
(\Omega_{ik}\mu_k-\mu_i).}
\label{eq:directed_edge_current}
\end{equation}
The force on the sender generated by this same directed edge is not a second copy of $J_{k\to i}$; it is the cotangent recoil $-\Omega_{ik}^{\top}J_{k\to i}$.  Consequently the complete local balance is
\begin{align}
\dot\pi_i^\mu
&+(\boldsymbol\Gamma\dot{\boldsymbol\mu})_i
+\alpha_i\bar\Lambda_{p_i}(\mu_i-\bar\mu_i)
+\Lambda_{o_i}(\mu_i-o_i)\nonumber\\
&=\sum_kJ_{k\to i}
-\sum_j\Omega_{ji}^{\top}J_{i\to j}
+f_{i,\mathrm{ext}}.
\label{eq:complete_current_balance}
\end{align}
Integrating Eq.~\eqref{eq:complete_current_balance} gives the momentum change over a declared interval.  It does not reduce to the incoming-current sum unless sender variables have explicitly been detached.


\subsubsection{Conservation and Non-Conservation}

Fixed asymmetric attention does not by itself destroy conservativity.  With the weights frozen, the sum of all directed edge energies is still one scalar potential, and Eqs.~\eqref{eq:edge_mean_gradients} and \eqref{eq:complete_current_balance} retain both receiver and sender derivatives.  Nonconservative behavior arises only after a declared truncation, such as treating each sender as detached and applying only the receiver current, or from explicit time dependence, damping, or external forcing.  This distinction parallels the systems-and-control separation between a Hamiltonian storage function and its dissipative ports \citep{XueHircheCao2020}.

Momentum conservation requires a continuous symmetry, not symmetric weights alone.  Let $a=(a_1,\ldots,a_N)$ be a covariantly constant translation generator satisfying $a_i=\Omega_{ik}a_k$ on every active edge.  If priors and observations are absent or translated with the state, $\boldsymbol\Gamma=0$, $f_{\mathrm{ext}}=0$, and the kinetic metric shares this symmetry, Noether's theorem gives
\begin{equation}
\boxed{
Q_a=\sum_i a_i^\top\pi_i^\mu,
\qquad
\frac{dQ_a}{dt}=0.}
\label{eq:covariant_momentum_conservation}
\end{equation}
In a flat common frame, $\Omega_{ik}=I$ and $a_i=a$, so Eq.~\eqref{eq:covariant_momentum_conservation} reduces componentwise to $d(\sum_i\pi_i^\mu)/dt=0$.  Anchoring terms, damping, external forcing, or the receiver-only detached truncation supply the exact nonconservation terms in Eq.~\eqref{eq:complete_current_balance}.  Attention asymmetry alone does not.

Finally, a loop holonomy $W_\gamma=\Omega_{ij}\Omega_{jk}\cdots\Omega_{\ell i}$ transforms by conjugation, $W_\gamma\mapsto g_iW_\gamma g_i^{-1}$.  Its conjugacy data---for example $\operatorname{tr}(W_\gamma^r)$ and characteristic-polynomial coefficients---are gauge invariant.  Wilson traces are the conventional compact-group observables.  By contrast, $\|W_\gamma-I\|_F$ is invariant under orthogonal or unitary conjugation but is frame dependent under a general $\mathrm{GL}(d)$ gauge transformation, so it cannot serve as a general gauge-invariant curvature magnitude.

%------------------------------------------------------------------------------
\subsection{Summary}
%------------------------------------------------------------------------------

\begin{table}[ht]
\centering
\caption{Conditional kinetic predictions and the quantities held fixed.  Every row assumes locally constant, independently specified positive $m$ and $k$ where a restoring stiffness is present.}
\label{tab:predictions}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{} p{3cm} p{5cm} p{5cm} @{}}
\toprule
\textbf{Protocol} & \textbf{Conditional formula} & \textbf{Controlled quantities} \\
\midrule
Harmonic overshoot & $d_{\mathrm{over}}=|v_0|\sqrt{m/k}$ & Initial velocity, $k$, damping protocol \\[6pt]
Ballistic stopping & $d_{\mathrm{stop}}=mv_0^2/(2f_0)$ & Initial velocity and opposing force \\[6pt]
Free oscillation & $\omega_d^2=k/m-\gamma^2/(4m^2)$ & $k$ and damping coefficient \\[6pt]
Forced resonance & $\omega_{\mathrm{res}}^2=k/m-\gamma^2/(2m^2)$ & Force amplitude, $k$, and damping coefficient \\[6pt]
Underdamped envelope & $\tau_{\mathrm{env}}=2m/\gamma$ & Damping coefficient \\[6pt]
Force-free coast & $\tau_{\mathrm{coast}}=m/\gamma$ & Damping coefficient \\
\bottomrule
\end{tabular}
\end{table}

These formulas characterize the conditional kinetic extension, not the primary Fisher natural-gradient flow.  Their empirical content lies in interventions that vary one operationally defined ingredient while holding the listed quantities fixed; replacing both $m$ and $k$ by the same VFE Hessian removes the claimed spectral dependence.

%==============================================================================
\subsection{Analytic Predictions and Falsification Tests}
\label{sec:computational_validation}
%==============================================================================

The preceding equations provide analytic predictions, not validation.  A useful test must estimate the kinetic metric $\mathbf M$ and restoring stiffness $\mathbf K$ through different interventions.  One protocol can estimate $\mathbf K$ from small static displacements around equilibrium, using the local slope of the restoring force while the pre-intervention state is held fixed.  A separate impulse or free-coast protocol can then estimate $\mathbf M$ and $\boldsymbol\Gamma$ from acceleration and velocity decay without reusing that static slope as mass.  Only after these measurements may Eqs.~\eqref{eq:conditional_damped_frequency} and \eqref{eq:conditional_resonance_frequency} be treated as parameter-free predictions for an independently driven trial.

The kinetic hypothesis is falsified if the independently measured tensors do not predict the observed modal frequencies, damping boundary, overshoot, or fixed-force response within prespecified uncertainty.  In particular, observing nonmonotonic motion is insufficient: Sampson, Restrepo, and Porter obtain oscillatory and excitable opinion dynamics without the kinetic mechanism posited here \citep{SampsonPorterRestrepo2025}.  A matched first-order feedback model, a time-varying-network model, and the conditional second-order model must therefore be compared on held-out trajectories under the same forcing schedule.  The kinetic account gains mechanism-specific support only if its independently estimated $\mathbf M$, $\mathbf K$, and $\boldsymbol\Gamma$ predict the intervention response better than those alternatives.

The social reduction below supplies a separate falsification target.  With fixed reversible weights and no anchors, the predicted decay rates are $\eta_\mu\lambda_s\ell_k$, where $\ell_k$ is a nonzero eigenvalue of the reversible graph Laplacian.  In the stated unanchored, symmetric reciprocal two-cluster reduction, persistent separation under a connected, positive, finite-temperature kernel would contradict the purely attractive model.  Conversely, persistence accompanied by a measured hard support break, a persistent anchor, signed influence, or active sampling would test an extended mechanism rather than the theorem proved here.

%==============================================================================
\section{Restricted Social-Dynamics Reductions}
\label{sec:classical_limits}
%==============================================================================

The social reductions use the primary Fisher--Rao flow, not Hessian-preconditioned descent and not an overdamped limit of the optional kinetic theory.  Row-normalized attention specifies relative allocation among candidate sources, so we introduce an unnormalized social strength $\lambda_s>0$ outside the attention row.  This parameter changes the rate of social relaxation without being erased by row normalization.

\subsection{Symmetric and Reversible DeGroot Flow}

\begin{proposition}[Reversible continuous-time DeGroot reduction]
\label{prop:reversible_degroot}
Let $W=[w_{ij}]$ be a fixed row-stochastic matrix that is reversible with respect to a strictly positive stationary distribution $\rho$, so $\rho_iw_{ij}=\rho_jw_{ji}$.  Assume flat transport, fixed common covariance $\Sigma_i=\sigma^2I$, no prior or sensory anchors, and fixed attention $\beta_{ij}=w_{ij}$.  Define the reciprocal relational loss
\begin{equation}
\mathcal F_{\mathrm{DG}}
=\frac{\lambda_s}{2}\sum_i\rho_i\sum_jw_{ij}\KL(q_i\|q_j)
=\frac{\lambda_s}{4\sigma^2}
\sum_{i,j}\rho_iw_{ij}\|\mu_i-\mu_j\|^2,
\label{eq:reversible_dirichlet_energy}
\end{equation}
and use the $\rho$-weighted product Fisher metric
$\mathbf G_\rho=\sigma^{-2}(D_\rho\otimes I)$.  Its Fisher--Rao natural-gradient flow is
\begin{equation}
\boxed{\dot{\boldsymbol\mu}
=-\eta_\mu\lambda_s[(I-W)\otimes I]\boldsymbol\mu.}
\label{eq:reversible_degroot_flow}
\end{equation}
For symmetric row-stochastic $W$, $\rho$ is uniform and the weighting contributes only an irrelevant global scale.  Forward Euler with $h\eta_\mu\lambda_s=1$ gives the discrete DeGroot step $\boldsymbol\mu^{n+1}=W\boldsymbol\mu^n$.
\end{proposition}

\begin{proof}
Detailed balance makes $c_{ij}=\rho_iw_{ij}$ symmetric.  Differentiating Eq.~\eqref{eq:reversible_dirichlet_energy} therefore gives
\begin{equation}
\nabla_{\mu_i}\mathcal F_{\mathrm{DG}}
=\frac{\lambda_s\rho_i}{\sigma^2}
\left(\mu_i-\sum_jw_{ij}\mu_j\right).
\end{equation}
The inverse $i$th Fisher block is $\sigma^2\rho_i^{-1}I$.  Substitution into the primary flow cancels both $\sigma^{-2}$ and $\rho_i$, yielding Eq.~\eqref{eq:reversible_degroot_flow}.  The Euler statement follows directly.
\end{proof}

This proposition is a subclass result.  A general directed row-stochastic matrix need not be reversible and need not arise as the complete gradient of this reciprocal scalar energy.  Dropping the sender derivative would manufacture a receiver-only directed update, but that is the detached truncation excluded in Section~\ref{sec:epistemic_status}, not a derivation of general DeGroot dynamics.

\subsection{Restricted Friedkin--Johnsen Equilibrium}

\begin{proposition}[Anchored reversible equilibrium]
\label{prop:restricted_fj}
Retain the assumptions of Proposition~\ref{prop:reversible_degroot}.  Let $u_i$ be a persistent prior mean and let $a_i\geq0$ be an agent-indexed anchor precision or coupling, with at least one anchor in every connected component.  Add
\begin{equation}
\mathcal F_{\mathrm{anchor}}
=\frac{1}{2\sigma^2}\sum_i\rho_i a_i\|\mu_i-u_i\|^2.
\label{eq:fj_anchor_energy}
\end{equation}
Writing $A=\operatorname{diag}(a_i)$, the unique equilibrium satisfies
\begin{equation}
\boxed{[A+\lambda_s(I-W)]\boldsymbol\mu^*=A\boldsymbol u,}
\label{eq:restricted_fj_linear_system}
\end{equation}
or, agentwise,
\begin{equation}
\mu_i^*=s_i u_i+(1-s_i)\sum_jw_{ij}\mu_j^*,
\qquad
s_i=\frac{a_i}{a_i+\lambda_s}.
\label{eq:restricted_fj_equilibrium}
\end{equation}
\end{proposition}

\begin{proof}
The equilibrium condition for $\mathcal F_{\mathrm{anchor}}+\mathcal F_{\mathrm{DG}}$ is
$a_i(\mu_i^*-u_i)+\lambda_s(\mu_i^*-\sum_jw_{ij}\mu_j^*)=0$ after cancelling the common factor $\rho_i/\sigma^2$.  Rearrangement gives Eqs.~\eqref{eq:restricted_fj_linear_system} and \eqref{eq:restricted_fj_equilibrium}.  The operator $A+\lambda_s(I-W)$ is self-adjoint and positive definite in the $D_\rho$-weighted inner product; equivalently, $D_\rho[A+\lambda_s(I-W)]$ is symmetric positive definite when every connected component contains an anchor.  The operator is therefore invertible, which gives uniqueness.
\end{proof}

Equation~\eqref{eq:restricted_fj_equilibrium} is an equilibrium correspondence, not a derivation of the general directed Friedkin--Johnsen iteration.  A common anchor precision produces only a common susceptibility.  Heterogeneous $s_i$ requires agent-indexed prior precision $a_i$, agent-indexed social coupling, or another explicitly heterogeneous mechanism; row normalization alone cannot produce it.

\subsection{Finite-Temperature Selectivity and Metastable Clustering}

For flat, isotropic beliefs the Gibbs rule is
\begin{equation}
\beta_{ij}
=\frac{\pi_{ij}\exp[-\|\mu_i-\mu_j\|^2/(2\sigma^2\tau)]}
{\sum_k\pi_{ik}\exp[-\|\mu_i-\mu_k\|^2/(2\sigma^2\tau)]}.
\label{eq:social_gibbs_kernel}
\end{equation}
It is similarity decreasing, conditional on the chosen kernel and candidate set.  If $\tau>0$ and every candidate prior $\pi_{ij}$ is positive, every candidate retains positive attention.  Similarity weighting therefore gives selective exposure, but it does not by itself remove cross-cluster attraction.

\begin{proposition}[Finite-temperature two-cluster contraction]
\label{prop:finite_temperature_contraction}
Consider an invariant two-cluster reduction with internally coincident means $a(t)$ and $b(t)$, common covariance $\sigma^2I$, flat transport, no anchors, and the symmetric reciprocal loss of Proposition~\ref{prop:reversible_degroot} with $\lambda_s>0$.  For either attention row, define the aggregate candidate-prior masses $\Pi_{\mathrm{in}}=\sum_{j\in\mathrm{same}}\pi_{ij}$ and $\Pi_{\mathrm{cross}}=\sum_{j\in\mathrm{other}}\pi_{ij}$.  Equal effective cluster priors means operationally that each row assigns equal aggregate prior mass to within-cluster and cross-cluster candidates, $\Pi_{\mathrm{in}}=\Pi_{\mathrm{cross}}>0$.  Before that specialization, the total cross-cluster attention is
\begin{equation}
p(d)=\left[1+\frac{\Pi_{\mathrm{in}}}{\Pi_{\mathrm{cross}}}
\exp\!\left(\frac{d^2}{2\sigma^2\tau}\right)\right]^{-1}.
\label{eq:unequal_cluster_cross_attention}
\end{equation}
Under equal effective cluster priors this becomes
\begin{equation}
p(d)=\frac{1}{1+\exp[d^2/(2\sigma^2\tau)]}>0,
\qquad d=\|b-a\|>0.
\label{eq:equal_cluster_cross_attention}
\end{equation}
The Fisher flow satisfies
\begin{equation}
\dot a=\eta_\mu\lambda_s p(d)(b-a),
\qquad
\dot b=\eta_\mu\lambda_s p(d)(a-b),
\end{equation}
and hence
\begin{equation}
\boxed{\dot d
=-2\eta_\mu\lambda_s p(d)d<0.}
\label{eq:strict_cluster_contraction}
\end{equation}
Thus a separated state is not an equilibrium of the positive finite-temperature attractive model.
\end{proposition}

\begin{proof}
The envelope identity in Eq.~\eqref{eq:reduced_vfe_envelope} permits differentiation at the optimized Gibbs weights without an additional first-order weight-response term.  Within-cluster residuals vanish.  Each remaining cross-cluster residual points from one cluster mean toward the other and has a strictly positive coefficient.  Subtracting the two mean equations gives
$\frac{d}{dt}(b-a)=-2\eta_\mu\lambda_s p(d)(b-a)$; differentiating its norm yields Eq.~\eqref{eq:strict_cluster_contraction}.
\end{proof}

The exact time to merge from $d_0$ to a declared tolerance $\delta\in(0,d_0)$ is
\begin{equation}
T_\delta
=\frac{1}{2\eta_\mu\lambda_s}
\left[
\log\frac{d_0}{\delta}
+\frac12\left\{\operatorname{Ei}\!\left(\frac{d_0^2}{2\sigma^2\tau}\right)
-\operatorname{Ei}\!\left(\frac{\delta^2}{2\sigma^2\tau}\right)\right\}
\right].
\label{eq:tolerance_merger_time}
\end{equation}
This time can be exponentially large when $d_0^2/(2\sigma^2\tau)$ is large, which justifies the term \emph{metastable}.  A tolerance $\varepsilon\in(0,1/2)$ for cross-cluster attention defines the boundary
\begin{equation}
d_\varepsilon^2
=2\sigma^2\tau\log\frac{1-\varepsilon}{\varepsilon}.
\label{eq:attention_tolerance_boundary}
\end{equation}
The earlier logarithmic rule is therefore a tolerance-defined merger-timescale boundary, not a stability or phase-transition threshold.  Taking $\varepsilon=1/N$ merely gives the special factor $\log(N-1)$.

Persistent polarization requires additional structure: hard support or severed network edges, persistent anchors with a separately proved separated equilibrium, signed or repulsive influence, or active confirmation-biased sampling that changes which evidence enters the candidate set.  None follows from positive finite-temperature attraction alone.


\subsection{Soft Bounded Confidence}

The Hegselmann--Krause rule uniformly averages agents inside a fixed confidence ball, while Deffuant dynamics update randomly selected pairs inside a hard bound \citep{hegselmann2002opinion,deffuant2000}.  Equation~\eqref{eq:social_gibbs_kernel} is related only at the level of graded selectivity: at finite temperature it assigns smoothly decreasing, generally nonzero weight as dissimilarity grows.  A reporting tolerance $\varepsilon$ can define an effective radius through Eq.~\eqref{eq:attention_tolerance_boundary}, but that radius depends on the chosen tolerance and is not a support boundary of the model.

The zero-temperature limit does not repair the difference.  Gibbs attention concentrates on the minimizing candidate or tied minimizers; it does not converge to a uniform average over a fixed Hegselmann--Krause confidence ball.  We therefore call the construction a finite-temperature, similarity-decreasing analog of bounded confidence, not an exact recovery of either classical rule.

\subsection{Confirmation Bias}

Within the present model, confirmation bias has one narrow correspondence: if the candidate set and similarity kernel in Eq.~\eqref{eq:social_gibbs_kernel} are imposed, consistent sources receive more attention than discrepant sources.  This is similarity-weighted selective exposure conditional on the kernel.  It does not derive active evidence seeking, biased likelihood evaluation, memory distortion, or asymmetric acceptance of equally diagnostic evidence.  Those mechanisms require an explicit sampling policy or observation model, as active-inference accounts of confirmation-biased policy selection make clear \citep{albarracin2022epistemic}.

\subsection{Social Impact Theory}

The relational force sums precision-weighted, attention-modulated source discrepancies, so source precision, similarity, and multiplicity admit a loose interpretation in the vocabulary of Social Impact Theory.  The correspondence is interpretive only.  Because each attention row is normalized, adding sources redistributes a fixed allocation unless the separate strength $\lambda_s$ is changed; the construction therefore does not derive Latan\'e's number law or its multiplicative strength--immediacy--number form.

\subsection{Diffusion of Innovations}

The present continuous belief state contains neither a binary adoption variable nor an adoption hazard.  It therefore yields no S-curve proposition and no adopter-category derivation.  A diffusion extension would need an explicit transition from belief to adoption and a population hazard with independently identified innovation and imitation channels, as in the Bass model \citep{Bass1969}.  Deriving that extension is future work rather than a consequence of the current attention dynamics.

\subsection{Proof-Status Summary}

\begin{table}[h]
\centering
\begin{tabular}{@{}p{3.1cm}p{3.2cm}p{6.1cm}@{}}
\toprule
\textbf{Claim} & \textbf{Status} & \textbf{Valid scope} \\
\midrule
DeGroot & Derived subclass & Fixed symmetric or reversible $W$, flat transport, common isotropic covariance, no anchors; general directed $W$ excluded. \\
Friedkin--Johnsen & Restricted equilibrium & Anchored reversible system; heterogeneous susceptibility requires agent-indexed anchor precision or coupling. \\
Bounded confidence & Soft analog & Finite-temperature similarity-decreasing attention; no hard support and no HK zero-temperature ball average. \\
Polarization & Metastability theorem & In the stated unanchored, symmetric reciprocal two-cluster reduction, positive finite-temperature attraction strictly contracts separation; persistence requires an added mechanism. \\
Confirmation bias & Conditional interpretation & Similarity-weighted selective exposure only. \\
Social Impact Theory & Interpretive only & No derivation of the number law under row normalization. \\
Diffusion of innovations & Not derived & Requires an adoption state and hazard model. \\
\bottomrule
\end{tabular}
\caption{Proof status of the social-dynamics claims.  The derived rows use the primary Fisher flow and do not depend on the kinetic postulate.}
\label{tab:rigor}
\end{table}


\section{Discussion}

\subsection{Unifying Phenomenological Models}

The strongest result is a gauge-covariant consensus objective with optimized entropy-retaining attention and a local four-part loss stiffness.  Its primary dynamics are Fisher--Rao natural-gradient flow.  In the flat, fixed-covariance, reversible subclass, that flow becomes continuous-time DeGroot consensus; persistent prior anchors give the restricted Friedkin--Johnsen equilibrium.  Finite-temperature Gibbs attention is a soft bounded-confidence analog.  In the stated unanchored, symmetric reciprocal two-cluster reduction, its positive attractive tail makes cluster separation metastable rather than stable.

The kinetic theory has a different status.  A positive local tensor $\mathbf M$ is an additional modeling choice, not a consequence of Fisher geometry or the loss Hessian.  When the same functional supplies both $\mathbf M$ and $\mathbf K=\mathbf H_F$, the spectrum is degenerate up to scale.  Nontrivial overshoot or resonance predictions therefore require independently operationalized $\mathbf M$, $\mathbf K$, and damping, as specified in Section~\ref{sec:computational_validation}.

\subsection{Interpretive Social and Psychological Mappings}

Belief perseverance is not derived from the one-state Fisher flow.  The force-free coast in Eq.~\eqref{eq:conditional_coasting_time} is a candidate kinetic mechanism, but explanatory perseverance after debunking generally requires a slow state that stores the explanation or causal model after the original evidence is removed.  A two-timescale extension must specify and estimate that state before the manuscript can claim a perseverance mechanism.

Confirmation bias is restricted here to similarity-weighted selective exposure conditional on Eq.~\eqref{eq:social_gibbs_kernel}.  The model does not establish active evidence search or asymmetric evaluation.  Likewise, the outgoing relational or recoil block measures contemporaneous local curvature generated by an agent's present source role.  It is not an accumulated memory of past followers, and it implies no theorem about leadership, empathy, morality, or inevitable rigidity.

Social Impact Theory remains an interpretive vocabulary, because normalized attention does not produce a number law.  Diffusion of innovations remains outside the state space until an adoption variable and hazard are added.  These boundaries are substantive: each missing mechanism requires a new state, policy, support rule, or signed interaction rather than a new rhetorical interpretation of the existing equations.


\subsection{Mechanism Identification}

Oscillation is not mechanism-identifying.  The conditional kinetic model, delayed feedback, time-dependent coupling, and individual--group feedback can all produce nonmonotonic trajectories; Sampson, Restrepo, and Porter provide a direct non-inertial example \citep{SampsonPorterRestrepo2025}.  Consequently an empirical study must perturb the restoring landscape and the kinetic response separately, estimate $\mathbf K$ and $\mathbf M$ from those distinct manipulations, and test the resulting out-of-sample frequency and decay predictions against matched first-order and feedback controls.  Merely observing overshoot, fitting a second-order curve, or recovering the formulas by integrating the same equations does not validate the kinetic postulate.

\subsection{Extensions Suggested by the Boundaries}

The exclusions above define concrete extensions.  Hard masks or endogenous network severing can produce disconnected supports; signed kernels can represent repulsion; active sampling policies can model confirmation-biased evidence acquisition; slow explanatory states can represent perseverance; and a Bass-type adoption hazard can connect continuous belief to diffusion.  Each extension should preserve the distinction between intrinsic Fisher geometry, objective curvature, and an optional kinetic metric, and each should be tested against the attractive finite-temperature baseline rather than attributed to it in advance.

\subsection{Gauge Curvature and Epistemic Frustration}
\label{sec:curvature_predictions}

The derivations in Section~\ref{sec:classical_limits} assumed flat gauge ($\Omega_{ij} = I$). It is important to recognize that the vertex-cocycle transport adopted in this paper, $\Omega_{ij}=U_iU_j^{-1}$, is pure gauge, and its holonomy around any closed loop $\gamma = (i \to j \to k \to i)$ collapses identically to the identity,
\begin{equation}
F_\gamma = \Omega_{ij}\Omega_{jk}\Omega_{ki}
=U_iU_j^{-1}U_jU_k^{-1}U_kU_i^{-1}=I,
\end{equation}
for every choice of frames $U_i\in\mathrm{GL}(d)$, because the connection is pure gauge and its curvature vanishes by the Maurer--Cartan identity. A restricted exponential-chart specialization may set $U_i=e^{\phi_i}$, giving $\Omega_{ij}=e^{\phi_i}e^{-\phi_j}$, but not every $\mathrm{GL}(d)$ frame lies in the image of the real matrix exponential. Nontrivial holonomy therefore does not arise from the present parameterization. It requires promoting the transport to an edge-local connection, $\Omega_{ij} = U_i\,e^{\delta_{ij}}\,U_j^{-1}$ with $\delta_{ij}$ an independent link variable valued in the Lie algebra (the lattice gauge-theory link variable), which reduces to the present flat case at $\delta_{ij} = 0$. Under that edge-relaxed extension the loop holonomy can satisfy $F_\gamma \neq I$, measuring the accumulated frame distortion when beliefs are transported around a social cycle, and nonzero curvature represents \emph{epistemic frustration}: the impossibility of consistently aligning all agents' reference frames simultaneously. The predictions outlined below are consequences of that edge-relaxed (non-pure-gauge) extension and are stated as a future-work program; they do not follow from the flat pure-gauge model studied in this paper.

\paragraph{Intransitive understanding.} Consider three agents---say a physicist ($i$), an economist ($j$), and a sociologist ($k$)---who each share overlapping vocabulary with their neighbors but encode concepts in incompatible frames. The physicist and economist may agree on what ``equilibrium'' means ($\Omega_{ij} \approx I$ in that subspace), and the economist and sociologist may agree on what ``rational agent'' means ($\Omega_{jk} \approx I$), but when the physicist and sociologist attempt direct communication, the composed transport $\Omega_{ij}\Omega_{jk}$ may not equal the direct transport $\Omega_{ik}$. The holonomy $F_\gamma = \Omega_{ij}\Omega_{jk}\Omega_{ki} \neq I$ quantifies this \emph{intransitivity of understanding}. The framework predicts that triadic closure in interdisciplinary networks should be inhibited by gauge curvature: agents who understand each other pairwise through intermediaries may nonetheless fail to reach mutual understanding directly, because the composed frame transformations accumulate distortion that a direct interaction would not.

\paragraph{Persistent disagreement under apparent consensus.} In a flat-gauge network, consensus ($\mu_i = \mu_j$ for all $i,j$) is an attainable equilibrium. With nonzero curvature, however, there may exist no global belief state that simultaneously minimizes all pairwise alignment costs $D_{\mathrm{KL}}(q_i \| \Omega_{ij} q_j)$. The social free energy develops a ``frustration energy'' analogous to spin glasses \citep{mezard1987spin}: the system cannot satisfy all local alignment constraints simultaneously, and the ground state exhibits residual disagreement even after exhaustive deliberation. This predicts that in networks with high gauge curvature (e.g., cross-cultural negotiations, interdisciplinary collaborations), consensus should be systematically harder to achieve than network topology alone would predict. The obstruction is geometric, not motivational.

\paragraph{Holonomy-induced polarization.} Even when two agents $i$ and $j$ hold identical beliefs in their respective frames ($\mu_i = \mu_j$ in each local coordinate system), the transported comparison $\Omega_{ij}\mu_j$ may differ from $\mu_i$ if the gauge frames are misaligned. Two communities may share factual beliefs but disagree because they encode those beliefs in rotated reference frames---for instance, framing the same economic policy as ``protecting workers'' versus ``restricting markets.'' The framework predicts that framing-induced polarization should be distinguishable from content-based polarization: agents with high mutual KL divergence $D_{\mathrm{KL}}(q_i \| \Omega_{ij}q_j)$ despite similar marginal beliefs $\mu_i \approx \mu_j$ are separated by gauge curvature rather than genuine disagreement. Interventions that ``flatten'' the gauge (e.g., establishing shared vocabulary or common framing) should reduce this component of polarization, whereas content-directed interventions should not.

\paragraph{Cultural translation losses.} Translating a concept from culture $A$ to culture $B$ to culture $C$ and back to $A$ should accumulate holonomy: the re-imported concept will generally differ from the original by a transformation $F_\gamma \in G$ that measures the total distortion. This predicts that in multilateral diplomacy or international scientific collaboration, the fidelity of concept transmission should degrade with the number of intermediate translations, and the degradation should depend on the curvature of the gauge connection (the degree of frame incompatibility) rather than on the semantic content of the beliefs themselves. Longitudinal studies of concept drift across translation chains could test this prediction quantitatively.

\paragraph{Curvature-driven instability.} In lattice gauge theory, high curvature configurations carry high energy and are dynamically suppressed. By analogy, social networks with high gauge curvature should be dynamically unstable: the frustrated free energy drives agents to either (a) reduce curvature by aligning their frames (cultural homogenization), (b) reduce connectivity by severing cross-group links (network fragmentation), or (c) reduce attention to frustrating neighbors (selective disengagement). A future phase diagram must use a declared conjugacy invariant, for example $\chi_\gamma=|\operatorname{tr}(F_\gamma)-d|$, together with the attention temperature $\kappa$. The Frobenius norm of $F_\gamma$ is not an admissible general-$\mathrm{GL}(d)$ observable because nonorthogonal conjugation changes it. This remains a proposed extension rather than a result of the present flat model.

\paragraph{Measurability.} While gauge frames $\phi_i$ are not directly observable, an edge-relaxed model could estimate conjugacy data of $F_\gamma$ from calibrated transport maps. Pairwise calibration followed by triadic composition would target Wilson traces or characteristic-polynomial coefficients; a raw Frobenius discrepancy $\|F_\gamma-I\|_F$ would be frame dependent under general $\mathrm{GL}(d)$. Whether such class functions correlate with triadic disagreement is a future falsification question, not a measurement established here.

\subsection{Limitations}

Several simplifying assumptions warrant future relaxation.

\textbf{Gaussian beliefs.} While analytically tractable, Gaussian distributions cannot capture the multi-modal posteriors characteristic of hypothesis competition, cognitive dissonance, or attitude ambivalence. Extension to exponential families is developed in Appendix~\ref{app:exponential_family}; extension to arbitrary distributions requires numerical methods or variational approximations.

\textbf{Restricted reductions.} The social theorems assume flat transport, common fixed covariance, and symmetric or reversible coupling.  General directed DeGroot or Friedkin--Johnsen dynamics, hard bounded confidence, persistent polarization, adoption diffusion, and active evidence selection require additional structure.  The two-cluster theorem does not classify every high-dimensional transient or every anchored equilibrium.

\textbf{Kinetic identification.} The manuscript supplies no empirical estimate of $\mathbf M$, $\mathbf K$, or $\boldsymbol\Gamma$.  The analytic oscillator formulas become testable only after these objects are independently operationalized, and oscillation alone cannot identify the kinetic mechanism.  The proposed matched controls and held-out predictions remain future work.



\section{Conclusion}

The primary contribution is a gauge-covariant population consensus energy with entropy-retaining optimized attention and a local four-part stiffness decomposition.  The induced first-order dynamics are Fisher--Rao natural-gradient flow.  Under explicit reversible, flat, fixed-covariance assumptions, that flow yields continuous-time DeGroot consensus and a restricted anchored Friedkin--Johnsen equilibrium.  Gibbs attention supplies soft similarity selectivity, while in the stated unanchored, symmetric reciprocal two-cluster reduction, finite-temperature positive attraction makes separated clusters metastable rather than stably polarized.

Belief inertia remains a conditional kinetic postulate.  The intrinsic Fisher metric $\mathbf G$, loss stiffness $\mathbf H_F$, and kinetic metric $\mathbf M$ are distinct, and choosing $\mathbf M=\mathbf H_F$ makes the same-functional oscillator spectrum trivial up to scale.  Overshoot, resonance, and coasting laws therefore become mechanism-specific predictions only when kinetic and restoring tensors are estimated independently.  Observed oscillation is not sufficient evidence for this mechanism.

The same calibration applies to the social interpretations.  Similarity-weighted attention can represent selective exposure, but it does not derive confirmation bias in general.  A force-free kinetic coast is a candidate component of perseverance, but an explanatory slow state is absent.  Recoil is local contemporaneous curvature, not accumulated leadership memory.  Social Impact Theory is interpretive, and diffusion awaits an adoption hazard.  These boundaries leave a narrower theory, but they also define the exact extensions and falsification tests needed to strengthen it.

\subsection*{Data Availability}
No new empirical data are reported or analyzed in this theoretical study.


\subsection*{Acknowledgments}
Claude was used for organizational and manuscript-clarity assistance.  The author takes responsibility for the mathematical statements and declares no funding or conflicts of interest.

\appendix

%==============================================================================
% APPENDIX B: GAUGE FRAME VARIATIONS AND PULLBACK GEOMETRY
%==============================================================================

\section{Gauge Frame Variations and Pullback Geometry}
\label{app:gauge}

The Hamiltonian formulation of belief dynamics reflects deep geometric structure. Each agent's belief space carries a gauge freedom: the choice of coordinate frame in which beliefs are expressed. Physical quantities must be invariant under these gauge transformations, while the dynamics must be covariant. This appendix develops the transformation theory for the mass matrix, momenta, and Hamilton's equations under gauge frame variations.

%------------------------------------------------------------------------------
\subsection{Gauge Structure of Multi-Agent Belief Systems}
%------------------------------------------------------------------------------

\subsubsection{The Principal Bundle}

The geometric setting is a principal $G$-bundle $\pi: P \to \mathcal{C}$ where $\mathcal{C}$ denotes the base manifold encoding agent positions and social network topology, $G \subseteq \mathrm{GL}(d)$ is the gauge group acting on belief space, and the fiber $\pi^{-1}(c)$ over each point $c \in \mathcal{C}$ is the space of reference frames. The natural choice is $G = \mathrm{GL}(d)$ (the maximal group under which KL divergence is invariant), though metric-preserving subgroups such as $\mathrm{SO}(d)$ are appropriate when agents share a common notion of statistical distance. The results in this appendix hold for any closed subgroup $G \subseteq \mathrm{GL}(d)$.

Each agent $i$ located at $c_i \in \mathcal{C}$ expresses beliefs in a local frame. The \textbf{transport operator} $\Omega_{ik} \in G$ relates agent $k$'s frame to agent $i$'s frame.

\subsubsection{Gauge Transformations}

A \textbf{gauge transformation} is a smooth assignment of group elements to each agent

\begin{equation}
g: \{1, \ldots, N\} \to G, \quad i \mapsto g_i
\end{equation}

Under this transformation, belief parameters transform as:
\begin{align}
\mu_i &\mapsto \mu_i' = g_i \mu_i \\
\Sigma_i &\mapsto \Sigma_i' = g_i \Sigma_i g_i^T \\
\Lambda_{q_i} &\mapsto \Lambda_{q_i}' = (g_i \Sigma_i g_i^T)^{-1} = g_i^{-T} \Lambda_{q_i} g_i^{-1}
\end{align}
The precision transforms contragrediently to the covariance because $\Lambda_{q_i} = \Sigma_i^{-1}$; on the metric-preserving subgroup $\mathrm{SO}(d)$, where $g_i^{-T} = g_i$, this reduces to $g_i \Lambda_{q_i} g_i^T$.
The prior and observation models are expressed in the same local frame, so their precision matrices transform similarly: $\bar\Lambda_{p_i}'=g_i^{-T}\bar\Lambda_{p_i}g_i^{-1}$ and $\Lambda_{o_i}'=g_i^{-T}\Lambda_{o_i}g_i^{-1}$.

The transport operators transform as:
\begin{equation}
\Omega_{ik} \mapsto \Omega_{ik}' = g_i \Omega_{ik} g_k^{-1}
\end{equation}

This ensures that the transported belief $\tilde{q}_k = \Omega_{ik}\cdot q_k$ transforms consistently:
\begin{equation}
\tilde{\mu}_k' = g_i \tilde{\mu}_k, \quad \tilde{\Lambda}_{q_k}' = g_i^{-T} \tilde{\Lambda}_{q_k} g_i^{-1}
\end{equation}

%------------------------------------------------------------------------------
\subsection{Transformation of the Affine Stiffness Blocks}
%------------------------------------------------------------------------------

\subsubsection{Mean Sector}

Because the local gauge-frame changes are affine linear, the frozen-attention mean-stiffness blocks transform by congruence. This does not extend ordinary Hessian covariance to arbitrary nonlinear reparameterizations away from a critical point.

\paragraph{Diagonal blocks:}
\begin{align}
[\mathbf H_F^\mu]_{ii}' &= \alpha_i\bar{\Lambda}_{p_i}' + \Lambda_{o_i}' + \sum_k \beta_{ik}\tilde{\Lambda}_{q_k}' + \sum_j \beta_{ji}\Lambda_{q_i}' \nonumber\\
&= \alpha_i g_i^{-T}\bar{\Lambda}_{p_i} g_i^{-1} + g_i^{-T}\Lambda_{o_i}g_i^{-1} + \sum_k \beta_{ik} g_i^{-T}\tilde{\Lambda}_{q_k} g_i^{-1} + \sum_j \beta_{ji} g_i^{-T}\Lambda_{q_i} g_i^{-1} \nonumber\\
&= g_i^{-T} \left[\alpha_i\bar{\Lambda}_{p_i} + \Lambda_{o_i} + \sum_k \beta_{ik}\tilde{\Lambda}_{q_k} + \sum_j \beta_{ji}\Lambda_{q_i}\right] g_i^{-1} \nonumber\\
&= g_i^{-T} [\mathbf H_F^\mu]_{ii} g_i^{-1}
\end{align}

\paragraph{Off-diagonal blocks:}
With the off-diagonal form $[\mathbf H_F^\mu]_{ik}=-\beta_{ik}\Omega_{ik}^{-T}\Lambda_{q_k}-\beta_{ki}\Lambda_{q_i}\Omega_{ki}^{-1}$ from Eq.~\eqref{eq:mass_offdiagonal},
\begin{align}
[\mathbf H_F^\mu]_{ik}' &= -\beta_{ik}(\Omega_{ik}')^{-T}\Lambda_{q_k}' - \beta_{ki}\Lambda_{q_i}'(\Omega_{ki}')^{-1} \nonumber\\
&= -\beta_{ik}(g_i\Omega_{ik}g_k^{-1})^{-T}(g_k^{-T}\Lambda_{q_k} g_k^{-1}) - \beta_{ki}(g_i^{-T}\Lambda_{q_i} g_i^{-1})(g_k\Omega_{ki}g_i^{-1})^{-1} \nonumber\\
&= -\beta_{ik} g_i^{-T}\Omega_{ik}^{-T}\Lambda_{q_k} g_k^{-1} - \beta_{ki} g_i^{-T}\Lambda_{q_i}\Omega_{ki}^{-1} g_k^{-1} \nonumber\\
&= g_i^{-T} [\mathbf H_F^\mu]_{ik} g_k^{-1}
\end{align}

\paragraph{Block matrix form:}

Define the block-diagonal mean-coordinate gauge matrix:
\begin{equation}
\mathsf g=\operatorname{diag}(g_1,g_2,\ldots,g_N)\in G^N.
\end{equation}

Then the full mean-sector stiffness transforms as
\begin{equation}
\boxed{(\mathbf H_F^\mu)' = \mathsf g^{-T}\mathbf H_F^\mu\mathsf g^{-1}}.
\end{equation}

This is the affine congruence law contragredient to $\mu'=\mathsf g\mu$. On $\mathrm{SO}(d)$ it reduces to $\mathsf g\mathbf H_F^\mu\mathsf g^T$.

\subsubsection{Covariance Sector}

For full vectorization, $\operatorname{vec}(\Sigma_i')=(g_i\otimes g_i)\operatorname{vec}(\Sigma_i)$. The stacked covariance-coordinate Jacobian is therefore
\begin{equation}
\mathsf g_\Sigma
=\operatorname{diag}(g_1\otimes g_1,\ldots,g_N\otimes g_N).
\label{eq:stacked_covariance_jacobian}
\end{equation}
The locally matched covariance-stiffness block involves the same Kronecker factors. Using the contragredient precision law $\Lambda_{q_i}'=g_i^{-T}\Lambda_{q_i}g_i^{-1}$ gives
\begin{align}
[\mathbf H_F^\Sigma]_{ii}' &= \frac{1}{2}(\Lambda_{q_i}' \otimes \Lambda_{q_i}') \cdot \left(\alpha_i + \sum_k \beta_{ik} + \sum_j \beta_{ji}\right) \nonumber\\
&= \frac{1}{2}(g_i^{-T}\Lambda_{q_i} g_i^{-1} \otimes g_i^{-T}\Lambda_{q_i} g_i^{-1}) \cdot \left(\alpha_i + \sum_k \beta_{ik} + \sum_j \beta_{ji}\right) \nonumber\\
&= \frac{1}{2}(g_i^{-T} \otimes g_i^{-T})(\Lambda_{q_i} \otimes \Lambda_{q_i})(g_i^{-1} \otimes g_i^{-1}) \cdot \left(\alpha_i + \sum_k \beta_{ik} + \sum_j \beta_{ji}\right)
\end{align}

The stacked full-vectorization transformation law is
\begin{equation}
\boxed{(\mathbf H_F^\Sigma)'=\mathsf g_\Sigma^{-T}\mathbf H_F^\Sigma\mathsf g_\Sigma^{-1}.}
\label{eq:stacked_covariance_hessian_congruence}
\end{equation}

Equations~\eqref{eq:stacked_covariance_jacobian}--\eqref{eq:stacked_covariance_hessian_congruence} use full $\operatorname{vec}$ coordinates. For independent $\operatorname{vech}$ coordinates, let $D_d$ be the duplication matrix and $L_d=(D_d^\top D_d)^{-1}D_d^\top$ the elimination matrix. The $i$th covariance Jacobian is then $L_d(g_i\otimes g_i)D_d$, and the Hessian congruence uses the block-diagonal assembly of these duplication/elimination-matrix pullbacks rather than $\mathsf g_\Sigma$ directly.

\subsubsection{Cross Blocks}

In the same full-vectorization convention, the mean-covariance cross blocks transform as
\begin{equation}
\boxed{(\mathbf H_F^{\mu\Sigma})'=\mathsf g^{-T}\mathbf H_F^{\mu\Sigma}\mathsf g_\Sigma^{-1}.}
\label{eq:stacked_mixed_hessian_congruence}
\end{equation}

%------------------------------------------------------------------------------
\subsection{Transformation of Momenta}
%------------------------------------------------------------------------------

For Hamilton's equations to be covariant, momenta must transform contragrediently to positions.  With a coupled kinetic metric their definition is the full cotangent relation, not a per-agent diagonal product.

\subsubsection{Mean Momentum}

For the stacked mean coordinate,
\begin{equation}
\boldsymbol\pi^\mu=\mathbf M^\mu\dot{\boldsymbol\mu},
\qquad
\boxed{\pi_i^\mu=\sum_k\mathbf M_{ik}^\mu\dot\mu_k.}
\end{equation}
Only a declared block-diagonal approximation gives $\pi_i^\mu=M_i\dot\mu_i$.  The mean momentum transforms as a covector, contragrediently to the position $\mu'=g\mu$:
\begin{equation}
\boxed{(\pi_i^\mu)' = g_i^{-T} \pi_i^\mu}
\end{equation}

This ensures the pairing $\langle\pi^\mu, \dot{\mu}\rangle$ is gauge-invariant, with velocities $\dot{\mu}' = g_i\dot{\mu}_i$ transforming as vectors:
\begin{equation}
\langle(\pi^\mu)', \dot{\mu}'\rangle = (g_i^{-T}\pi_i^\mu)^T(g_i\dot{\mu}_i) = (\pi_i^\mu)^T g_i^{-1} g_i \dot{\mu}_i = (\pi_i^\mu)^T\dot{\mu}_i = \langle\pi^\mu, \dot{\mu}\rangle
\end{equation}

\subsubsection{Covariance Momentum}

The covariance momentum $\Pi^\Sigma \in \mathrm{Sym}(d)$ transforms contragrediently to $\Sigma' = g_i\Sigma_i g_i^T$:
\begin{equation}
\boxed{(\Pi_i^\Sigma)' = g_i^{-T} \Pi_i^\Sigma g_i^{-1}}
\end{equation}

The pairing with $\dot{\Sigma}$ uses the trace:
\begin{equation}
\mathrm{tr}[(\Pi^\Sigma)'\dot{\Sigma}'] = \mathrm{tr}[(g_i^{-T}\Pi_i^\Sigma g_i^{-1})(g_i\dot{\Sigma}_i g_i^T)] = \mathrm{tr}[\Pi_i^\Sigma\dot{\Sigma}_i]
\end{equation}
where we used cyclicity of the trace and $g_i^{-1} g_i = I$, valid for any $g_i \in \mathrm{GL}(d)$.

%------------------------------------------------------------------------------
\subsection{Covariance of Hamilton's Equations}
%------------------------------------------------------------------------------

\subsubsection{Velocity Equation}

The velocity equation $\dot{\mu} = (\mathbf{M}^\mu)^{-1}\pi^\mu$ transforms as:
\begin{align}
\dot{\mu}' &= ((\mathbf{M}^\mu)')^{-1}(\pi^\mu)' \nonumber\\
&= (\mathsf g^{-T}\mathbf{M}^\mu\mathsf g^{-1})^{-1}\mathsf g^{-T}\pi^\mu \nonumber\\
&= \mathsf g(\mathbf{M}^\mu)^{-1}\mathsf g^{T}\mathsf g^{-T}\pi^\mu \nonumber\\
&= \mathsf g(\mathbf{M}^\mu)^{-1}\pi^\mu = \mathsf g\dot{\mu}
\end{align}

Velocities therefore transform as vectors, $\dot{\mu}'=\mathsf g\dot{\mu}$, consistently with $\mu'=\mathsf g\mu$, for any $G\subseteq\mathrm{GL}(d)$. If a mean-sector kinetic block $\mathbf M^\mu$ is separately chosen with $(\mathbf M^\mu)'=\mathsf g^{-T}\mathbf M^\mu\mathsf g^{-1}$, then its quadratic kinetic term is gauge invariant. The full-state kinetic metric requires the larger Jacobian stated in the summary below.

\subsubsection{Force Equation}

The force equation involves the free energy gradient. Since $F$ is gauge-invariant and $\mu' = g_i\mu_i$, the gradient transforms as a covector under the chain rule:
\begin{equation}
\left(\frac{\partial F}{\partial\mu_i}\right)' = g_i^{-T} \frac{\partial F}{\partial\mu_i}
\end{equation}

This follows from the chain rule and the invariance of $F$ under gauge transformations when transport operators transform consistently.

The geodesic force transforms the same way, ensuring full covariance with the momentum law $(\pi^\mu)'=\mathsf g^{-T}\pi^\mu$:
\begin{equation}
\boxed{\dot{\pi}'=\mathsf g^{-T}\dot{\pi}}
\end{equation}

%------------------------------------------------------------------------------
\subsection{The Connection and Its Variation}
%------------------------------------------------------------------------------

\subsubsection{Connection 1-Form}

The transport operators $\Omega_{ik}$ encode a discrete connection on the agent network. For agents connected along an edge $e = (i,k)$, define:
\begin{equation}
A_e = \Omega_{ik} \in G
\end{equation}

Under gauge transformation:
\begin{equation}
A_e \mapsto A_e' = g_i A_e g_k^{-1}
\end{equation}

This is the discrete analog of the gauge transformation $A \mapsto gAg^{-1} + g dg^{-1}$ for continuous connections.

\subsubsection{Curvature}

The discrete loop holonomy around $\gamma=(i\to j\to k\to i)$ is
\begin{equation}
F_\gamma = \Omega_{ij}\Omega_{jk}\Omega_{ki} \in G
\end{equation}

This is gauge-covariant: $F_\gamma'=g_iF_\gamma g_i^{-1}$.  Accordingly, the gauge-invariant data are conjugacy invariants of $F_\gamma$, not its coordinate entries.

A \textbf{flat connection} satisfies $F_\gamma = I$ for all loops, meaning beliefs can be consistently parallel-transported around any cycle. Nonzero curvature represents "information geometry frustration" whereby belief frames cannot be consistently aligned around cycles.

\subsubsection{Variation of Connection}

Consider an infinitesimal variation of the connection:
\begin{equation}
\delta\Omega_{ik} = \omega_{ik}   \Omega_{ik}, \quad \omega_{ik} \in \mathfrak{g}
\end{equation}

The variation of transported precision, using $\tilde{\Lambda}_k = \Omega_{ik}^{-T}\Lambda_k\Omega_{ik}^{-1}$ and $\delta(\Omega^{-1}) = -\Omega^{-1}(\delta\Omega)\Omega^{-1}$, is
\begin{equation}
\delta\tilde{\Lambda}_k = -\omega_{ik}^T\tilde{\Lambda}_k - \tilde{\Lambda}_k\omega_{ik}.
\end{equation}
For $G = \mathrm{SO}(d)$, the antisymmetry $\omega^T = -\omega$ reduces this to the matrix commutator $\delta\tilde{\Lambda}_k = \omega_{ik}\tilde{\Lambda}_k - \tilde{\Lambda}_k\omega_{ik} = [\omega_{ik}, \tilde{\Lambda}_k]$.

%------------------------------------------------------------------------------
\subsection{Variation of the Loss Stiffness Under Connection Changes}
%------------------------------------------------------------------------------

\subsubsection{Diagonal Block Variation}

\begin{equation}
\delta[\mathbf H_F^\mu]_{ii} = \sum_k \beta_{ik}\,\delta\tilde{\Lambda}_{q_k} = \sum_k \beta_{ik}[\omega_{ik},\tilde{\Lambda}_{q_k}]
\end{equation}

\subsubsection{Off-Diagonal Block Variation}

We specialize to $G=\mathrm{SO}(d)$, where the off-diagonal stiffness block is $[\mathbf H_F^\mu]_{ik}=-\beta_{ik}\Omega_{ik}\Lambda_{q_k}-\beta_{ki}\Lambda_{q_i}\Omega_{ki}^T$ (for general $\mathrm{GL}(d)$ the transposes are replaced by inverses, as in Eq.~\eqref{eq:mass_offdiagonal}):
\begin{align}
\delta[\mathbf H_F^\mu]_{ik} &= -\beta_{ik}\,\delta(\Omega_{ik}\Lambda_{q_k}) - \beta_{ki}\,\delta(\Lambda_{q_i}\Omega_{ki}^T) \nonumber\\
&= -\beta_{ik}\omega_{ik}\Omega_{ik}\Lambda_{q_k} - \beta_{ki}\Lambda_{q_i}(\omega_{ki}\Omega_{ki})^T \nonumber\\
&= -\beta_{ik}\omega_{ik}\Omega_{ik}\Lambda_{q_k} - \beta_{ki}\Lambda_{q_i}\Omega_{ki}^T\omega_{ki}^T
\end{align}

For the $\mathrm{SO}(d)$ case, using $\omega_{ki}^T = -\omega_{ki}$:
\begin{equation}
\boxed{\delta[\mathbf H_F^\mu]_{ik}=-\beta_{ik}\omega_{ik}\Omega_{ik}\Lambda_{q_k}+\beta_{ki}\Lambda_{q_i}\Omega_{ki}^T\omega_{ki}}
\end{equation}

%------------------------------------------------------------------------------
\subsection{Pullback Geometry}
%------------------------------------------------------------------------------

The \textbf{pullback} of the metric under a map $\phi: \mathcal{Q} \to \mathcal{Q}$ is central to understanding how geometry transforms under coordinate changes or symmetry actions.

\subsubsection{Pullback of the Fisher-Rao Metric}

Let $\phi_g: \mathcal{Q} \to \mathcal{Q}$ be the action of gauge transformation $g$:
\begin{equation}
\phi_g(\mu, \Sigma) = (g\mu, g\Sigma g^T)
\end{equation}

The pullback metric is:
\begin{equation}
(\phi_g^*\mathcal{G})_{(\mu,\Sigma)}(v, w) = \mathcal{G}_{\phi_g(\mu,\Sigma)}(d\phi_g \cdot v, d\phi_g \cdot w)
\end{equation}

For the Fisher-Rao metric, gauge invariance implies:
\begin{equation}
\boxed{\phi_g^*\mathcal{G} = \mathcal{G}}
\end{equation}

The metric is gauge-invariant representing the geometric content of our transformation laws.

\subsubsection{Horizontal and Vertical Decomposition}

The tangent space at each point decomposes as:
\begin{equation}
T_{(\mu,\Sigma)}\mathcal{Q} = H_{(\mu,\Sigma)} \oplus V_{(\mu,\Sigma)}
\end{equation}

The tangent space at each point splits into a \textbf{vertical space} $V$ consisting of directions along gauge orbits (pure gauge changes) and a \textbf{horizontal space} $H$ consisting of directions orthogonal to gauge orbits (physical changes). The connection determines the horizontal subspace. A vector $v = (\delta\mu, \delta\Sigma)$ is horizontal if:
\begin{equation}
\mathcal{G}(v, \xi_X) = 0 \quad \forall X \in \mathfrak{g}
\end{equation}
where $\xi_X$ is the vector field generated by $X$.

\subsubsection{Gauge-Invariant Quantities}

Only horizontal components of velocities and momenta correspond to observables. Gauge-invariant quantities include the \textbf{consensus divergence} $\|\mu_i - \tilde{\mu}_k\|_{\tilde{\Lambda}_{q_k}}^2 = (\mu_i - \tilde{\mu}_k)^T\tilde{\Lambda}_{q_k}(\mu_i - \tilde{\mu}_k)$, the \textbf{free energy} $F[\{q_i\}]$, the conditional \textbf{Hamiltonian} $H=\frac12\langle\pi,\mathbf M^{-1}\pi\rangle+F$, and the \textbf{inter-agent KL divergence} $\mathrm{KL}(q_i\|\Omega_{ik}\cdot q_k)$. Each scalar is invariant when beliefs, transports, and the postulated kinetic metric transform consistently. Loop holonomy transforms by conjugation, $F_\gamma\to g_iF_\gamma g_i^{-1}$, so $\operatorname{tr}(F_\gamma^r)$, the characteristic polynomial, and the eigenvalue multiset are gauge invariant. Wilson traces are the standard compact-group observables. The Frobenius norm $\|F_\gamma-I\|_F$ is invariant under orthogonal or unitary conjugation but is frame dependent for general $\mathrm{GL}(d)$ transformations.


%------------------------------------------------------------------------------
\subsection{Summary: Gauge-Covariant Hamiltonian Mechanics}
%------------------------------------------------------------------------------

\begin{tcolorbox}[title=Gauge Transformation Laws]
\textbf{Positions:}
\begin{align}
\mu_i' &= g_i\mu_i & \Sigma_i' &= g_i\Sigma_i g_i^T
\end{align}

\textbf{Momenta (contragredient to positions):}
\begin{align}
(\pi_i^\mu)' &= g_i^{-T}\pi_i^\mu & (\Pi_i^\Sigma)' &= g_i^{-T}\Pi_i^\Sigma g_i^{-1}
\end{align}

\textbf{Full-state kinetic metric (a $(0,2)$-tensor):}
\begin{equation}
J=\operatorname{diag}(\mathsf g,\mathsf g_\Sigma),
\qquad
\boxed{\mathbf M'=J^{-T}\mathbf M J^{-1}.}
\label{eq:full_state_kinetic_congruence}
\end{equation}
Here $J$ uses full $\operatorname{vec}(\Sigma_i)$ covariance coordinates. In $\operatorname{vech}$ coordinates, its covariance block must be replaced by the duplication/elimination-matrix Jacobian described after Eq.~\eqref{eq:stacked_covariance_hessian_congruence}. On $\mathrm{SO}(d)$ these transformations reduce to the corresponding orthogonal congruences.

\textbf{Transport Operators:}
\begin{equation}
\Omega_{ik}' = g_i\Omega_{ik}g_k^{-1}
\end{equation}

\textbf{Hamilton's Equations:} Fully covariant under these transformations.

\textbf{Observables:} Gauge-invariant quantities include $F$, the conditional $H$, all inter-agent divergences, and conjugacy invariants of loop holonomy.  A general-$\mathrm{GL}(d)$ Frobenius holonomy norm is frame dependent.
\end{tcolorbox}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix: Hamiltonian Mechanics on Statistical Manifolds
% With Explicit Sensory Likelihood
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Gaussian Loss Stiffness and Conditional Mechanics}
\label{app:hamiltonian}

This appendix records the detailed first and second derivatives of the Gaussian objective with explicit sensory evidence. The second variations define the affine-coordinate loss stiffness $\mathbf H_F$; they do not, by themselves, define the positive kinetic metric $\mathbf M$. We work in the quasi-static approximation where prior parameters $(\bar{\mu}_i, \bar{\Sigma}_i)$ evolve slowly relative to beliefs $(\mu_i, \Sigma_i)$.

%-----------------------------------------------------------------------
\subsection{Setup and Notation}
%-----------------------------------------------------------------------

Each agent $i$ maintains a belief distribution $q_i = \mathcal{N}(\mu_i, \Sigma_i)$ anchored to a fixed prior $p_i = \mathcal{N}(\bar{\mu}_i, \bar{\Sigma}_i)$ and receives observations $o_i$ through a likelihood $p(o_i \mid \theta) = \mathcal{N}(o_i; \theta, \Sigma_{o_i})$. Define:
%
\begin{align}
\Lambda_{q_i} &= \Sigma_i^{-1} & &\text{(belief precision)} \\
\bar{\Lambda}_{p_i} &= \bar{\Sigma}_i^{-1} & &\text{(prior precision)} \\
\Lambda_{o_i} &= \Sigma_{o_i}^{-1} & &\text{(observation precision)} \\
\tilde{\mu}_k &= \Omega_{ik}\mu_k & &\text{(transported mean)} \\
\tilde{\Lambda}_{q_k} &= \left(\Omega_{ik}\Sigma_k\Omega_{ik}^T\right)^{-1} = \Omega_{ik}^{-T}\Lambda_{q_k}\Omega_{ik}^{-1} & &\text{(transported precision)}
\end{align}
%
where $\Omega_{ik}\in G\subseteq\mathrm{GL}(d)$ is the gauge transport operator from agent $k$'s frame to agent $i$'s frame, given canonically by $\Omega_{ik}=U_iU_k^{-1}$. In the restricted exponential-chart specialization $U_i=e^{\phi_i}$ with $\phi_i\in\mathfrak g$, this becomes $\Omega_{ik}=e^{\phi_i}e^{-\phi_k}$; this chart does not represent every $\mathrm{GL}(d)$ frame.

%-----------------------------------------------------------------------
\subsection{The Extended Free Energy Functional}
%-----------------------------------------------------------------------

The complete scalar is Eq.~\eqref{eq:canonical_full_vfe}; this appendix does not define a second population objective. For the fixed-attention derivatives below, the entropy term is retained in $\mathcal F_{\mathrm{full}}$ but contributes no Gaussian-coordinate derivative. We write $\mathcal F=\mathcal F_{\mathrm{full}}$ for compactness and allow the explicit adaptive-prior replacement in Eq.~\eqref{eq:adaptive_self_coupling}. Accordingly, $\alpha_i=1$ denotes the canonical prior sector, while a fixed adaptive value carries the factor $\alpha_i$ through both belief-coordinate gradients and Hessian blocks. If $\alpha_i$ has already been optimized, envelope derivatives instead use $\alpha_i^*$ as specified in Eqs.~\eqref{eq:adaptive_envelope_derivative}--\eqref{eq:adaptive_prior_covariance_gradient}.

%-----------------------------------------------------------------------
\subsection{Component Free Energies for Gaussians}
%-----------------------------------------------------------------------

\subsubsection{KL Divergence Between Gaussians}

For $q = \mathcal{N}(\mu_q, \Sigma_q)$ and $p = \mathcal{N}(\mu_p, \Sigma_p)$:
%
\begin{equation}
D_{\mathrm{KL}}(q \| p) = \frac{1}{2}\left[\mathrm{tr}(\Sigma_p^{-1}\Sigma_q) + (\mu_p - \mu_q)^T\Sigma_p^{-1}(\mu_p - \mu_q) - d + \ln\frac{|\Sigma_p|}{|\Sigma_q|}\right]
\label{eq:kl_gaussian}
\end{equation}

\subsubsection{Expected Log-Likelihood}

For the Gaussian likelihood $p(o_i \mid \theta) = \mathcal{N}(o_i; \theta, \Sigma_{o_i})$:
%
\begin{align}
\mathbb{E}_{q_i}[\log p(o_i \mid \theta)] &= -\frac{d}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{o_i}| - \frac{1}{2}\mathbb{E}_{q_i}\left[(o_i - \theta)^T\Lambda_{o_i}(o_i - \theta)\right]
\end{align}
%
The quadratic expectation evaluates to:
%
\begin{equation}
\mathbb{E}_{q_i}\left[(o_i - \theta)^T\Lambda_{o_i}(o_i - \theta)\right] = (o_i - \mu_i)^T\Lambda_{o_i}(o_i - \mu_i) + \mathrm{tr}(\Lambda_{o_i}\Sigma_i)
\end{equation}
%
Therefore:
%
\begin{equation}
\boxed{
-\mathbb{E}_{q_i}[\log p(o_i \mid \theta)] = \frac{1}{2}(o_i - \mu_i)^T\Lambda_{o_i}(o_i - \mu_i) + \frac{1}{2}\mathrm{tr}(\Lambda_{o_i}\Sigma_i) + \mathrm{const}
}
\label{eq:neg_log_likelihood}
\end{equation}

%-----------------------------------------------------------------------
\subsection{First Variations (Gradient)}
%-----------------------------------------------------------------------

\subsubsection{Prior Term: $\alpha_iD_{\mathrm{KL}}(q_i \| p_i)$}

\begin{align}
\frac{\partial [\alpha_iD_{\mathrm{KL}}(q_i \| p_i)]}{\partial \mu_i} &= \alpha_i\bar{\Lambda}_{p_i}(\mu_i - \bar{\mu}_i) \label{eq:grad_prior_mu} \\[6pt]
\frac{\partial [\alpha_iD_{\mathrm{KL}}(q_i \| p_i)]}{\partial \Sigma_i} &= \frac{\alpha_i}{2}(\bar{\Lambda}_{p_i} - \Lambda_{q_i}) \label{eq:grad_prior_sigma}
\end{align}
Here $\alpha_i=1$ in the canonical sector, while the envelope-reduced adaptive sector uses $\alpha_i=\alpha_i^*$.

\subsubsection{Consensus Term: $D_{\mathrm{KL}}(q_i \| \tilde{q}_k)$}

With respect to receiver $i$:
\begin{align}
\frac{\partial D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \mu_i} &= \tilde{\Lambda}_{q_k}(\mu_i - \tilde{\mu}_k) \label{eq:grad_consensus_mu_i} \\[6pt]
\frac{\partial D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \Sigma_i} &= \frac{1}{2}(\tilde{\Lambda}_{q_k} - \Lambda_{q_i}) \label{eq:grad_consensus_sigma_i}
\end{align}

With respect to sender $k$:
\begin{align}
\frac{\partial D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \mu_k} &= \Lambda_{q_k}\Omega_{ik}^{-1}(\tilde{\mu}_k - \mu_i) \label{eq:grad_consensus_mu_k} \\[6pt]
\frac{\partial D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \Sigma_k} &= \frac{1}{2}\Omega_{ik}^T\left[\tilde{\Lambda}_{q_k} - \tilde{\Lambda}_{q_k}\Sigma_i\tilde{\Lambda}_{q_k}\right]\Omega_{ik} \label{eq:grad_consensus_sigma_k}
\end{align}

\subsubsection{Sensory Term: $-\mathbb{E}_{q_i}[\log p(o_i \mid \theta)]$}

\begin{align}
\frac{\partial}{\partial \mu_i}\left[-\mathbb{E}_{q_i}[\log p(o_i \mid \theta)]\right] &= \Lambda_{o_i}(\mu_i - o_i) \label{eq:grad_sensory_mu} \\[6pt]
\frac{\partial}{\partial \Sigma_i}\left[-\mathbb{E}_{q_i}[\log p(o_i \mid \theta)]\right] &= \frac{1}{2}\Lambda_{o_i} \label{eq:grad_sensory_sigma}
\end{align}

\subsubsection{Total Gradient}

\begin{equation}
\boxed{
\frac{\partial \mathcal{F}}{\partial \mu_i} = \alpha_i\bar{\Lambda}_{p_i}(\mu_i - \bar{\mu}_i) + \sum_k \beta_{ik}\tilde{\Lambda}_{q_k}(\mu_i - \tilde{\mu}_k) + \sum_j \beta_{ji}\Lambda_{q_i}\Omega_{ji}^{-1}(\tilde{\mu}_i^{(j)} - \mu_j) + \Lambda_{o_i}(\mu_i - o_i)
}
\label{eq:total_gradient_mu}
\end{equation}
%
where $\tilde{\mu}_i^{(j)} = \Omega_{ji}\mu_i$ is agent $i$'s mean transported into agent $j$'s frame.

\begin{equation}
\boxed{
\frac{\partial \mathcal{F}}{\partial \Sigma_i} = \frac{\alpha_i}{2}(\bar{\Lambda}_{p_i} - \Lambda_{q_i}) + \sum_k \frac{\beta_{ik}}{2}(\tilde{\Lambda}_{q_k} - \Lambda_{q_i}) + \sum_j \frac{\beta_{ji}}{2}\Omega_{ji}^T\left[\tilde{\Lambda}_{qi}^{(j)} - \tilde{\Lambda}_{qi}^{(j)}\Sigma_j\tilde{\Lambda}_{qi}^{(j)}\right]\Omega_{ji} + \frac{1}{2}\Lambda_{o_i}
}
\label{eq:total_gradient_sigma}
\end{equation}

%-----------------------------------------------------------------------
\subsection{Second Variations and the Loss Hessian}
%-----------------------------------------------------------------------

The coordinate Hessian $\mathbf H_F=\partial^2\mathcal F/\partial\xi\partial\xi^\top$ is the local stiffness of the selected objective. It is distinct from the intrinsic Fisher--Rao metric $\mathbf G$ and from any separately postulated kinetic metric $\mathbf M$.

\subsubsection{Mean Sector: $\partial^2\mathcal{F}/\partial\mu\partial\mu^T$}

\paragraph{Diagonal blocks $(i = k)$:}

From prior:
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| p_i)}{\partial \mu_i \partial \mu_i^T} = \bar{\Lambda}_{p_i}
\end{equation}

From consensus (as receiver):
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \mu_i \partial \mu_i^T} = \tilde{\Lambda}_{q_k} = \Omega_{ik}^{-T}\Lambda_{q_k}\Omega_{ik}^{-1} = \left(\Omega_{ik}\Sigma_k\Omega_{ik}^T\right)^{-1}
\end{equation}

From consensus (as sender to agent $j$):
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_j \| \tilde{q}_i)}{\partial \mu_i \partial \mu_i^T} = \Omega_{ji}^T\tilde{\Lambda}_{qi}^{(j)}\Omega_{ji} = \Lambda_{q_i}
\end{equation}

From sensory evidence:
\begin{equation}
\frac{\partial^2}{\partial \mu_i \partial \mu_i^T}\left[-\mathbb{E}_{q_i}[\log p(o_i \mid \theta)]\right] = \Lambda_{o_i}
\end{equation}

\textbf{Total diagonal stiffness:}
\begin{equation}
\boxed{
[\mathbf{H}_F^\mu]_{ii} = \underbrace{\alpha_i\bar{\Lambda}_{p_i}}_{\text{prior}} + \underbrace{\sum_k \beta_{ik}\tilde{\Lambda}_{q_k}}_{\text{incoming relational}} + \underbrace{\sum_j \beta_{ji}\Lambda_{q_i}}_{\text{outgoing recoil}} + \underbrace{\Lambda_{o_i}}_{\text{sensory}}
}
\label{eq:mass_diagonal}
\end{equation}

\paragraph{Off-diagonal blocks $(i \neq k)$:}

From $D_{\mathrm{KL}}(q_i \| \tilde{q}_k)$, using $\tilde{\Lambda}_{q_k} = \Omega_{ik}^{-T}\Lambda_{q_k}\Omega_{ik}^{-1}$ and the Jacobian $\partial\tilde{\mu}_k/\partial\mu_k = \Omega_{ik}$:
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \mu_i \partial \mu_k^T} = -\tilde{\Lambda}_{q_k}\Omega_{ik} = -\Omega_{ik}^{-T}\Lambda_{q_k}
\end{equation}

From $D_{\mathrm{KL}}(q_k \| \tilde{q}_i)$ (if $k$ also listens to $i$):
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_k \| \tilde{q}_i)}{\partial \mu_i \partial \mu_k^T} = -\Lambda_{q_i}\Omega_{ki}^{-1}
\end{equation}

The sensory term does not couple different agents. Therefore:
\begin{equation}
\boxed{
[\mathbf{H}_F^\mu]_{ik} = -\beta_{ik}\Omega_{ik}^{-T}\Lambda_{q_k} - \beta_{ki}\Lambda_{q_i}\Omega_{ki}^{-1} \quad (i \neq k)
}
\label{eq:mass_offdiagonal}
\end{equation}


By Schwarz's theorem, this Hessian is symmetric: using $\Lambda=\Lambda^T$ one verifies that $[\mathbf H_F^\mu]_{ik}=([\mathbf H_F^\mu]_{ki})^T$ for any frozen $\beta_{ik}$ and invertible $\Omega_{ik}$. The sender/recoil derivative is required because the scalar objective contains every directed edge on which agent $i$ appears; asymmetry of the frozen attention weights does not make this complete scalar potential nonconservative.


%-----------------------------------------------------------------------
\subsubsection{Covariance Sector: $\partial^2\mathcal{F}/\partial\Sigma\partial\Sigma$}
%-----------------------------------------------------------------------

For matrix-valued variables, we use the directional derivative convention:
\begin{equation}
\frac{\partial^2 f}{\partial \Sigma \partial \Sigma}[V, W] = \lim_{\epsilon \to 0} \frac{1}{\epsilon}\left(\left.\frac{\partial f}{\partial \Sigma}\right|_{\Sigma + \epsilon W} - \left.\frac{\partial f}{\partial \Sigma}\right|_{\Sigma}\right)[V]
\end{equation}

\paragraph{Key identity:}
\begin{equation}
\frac{\partial}{\partial \Sigma}(\Sigma^{-1}) = -\Sigma^{-1} \otimes \Sigma^{-1}
\end{equation}

\paragraph{Diagonal blocks $(i = k)$:}

From prior:
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| p_i)}{\partial \Sigma_i \partial \Sigma_i}[V, W] = \frac{1}{2}\mathrm{tr}\left[\Lambda_{q_i}V\Lambda_{q_i}W\right]
\end{equation}

In tensor notation:
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| p_i)}{\partial \Sigma_i \partial \Sigma_i} = \frac{1}{2}(\Lambda_{q_i} \otimes \Lambda_{q_i})
\end{equation}

At edgewise consensus, the receiver and sender covariance Hessians reduce to the same local quadratic form. The prior contribution carries the self-coupling weight $\alpha_i$, because its scalar term is $\alpha_i D_{\mathrm{KL}}(q_i\|p_i)$.

\textbf{Critical observation:} The sensory term $\frac{1}{2}\mathrm{tr}(\Lambda_{o_i}\Sigma_i)$ is \textit{linear} in $\Sigma_i$, so its second derivative \textbf{vanishes}:
\begin{equation}
\frac{\partial^2}{\partial \Sigma_i \partial \Sigma_i}\left[\mathrm{tr}(\Lambda_{o_i}\Sigma_i)\right] = 0
\end{equation}

Therefore:
\begin{equation}
\boxed{
[\mathbf{H}_F^\Sigma]_{ii} = \frac{1}{2}(\Lambda_{q_i} \otimes \Lambda_{q_i}) \cdot \left(\alpha_i + \sum_k \beta_{ik} + \sum_j \beta_{ji}\right)
}
\label{eq:mass_covariance}
\end{equation}

The sensory precision $\Lambda_{o_i}$ does \textbf{not} contribute to this covariance-sector stiffness.

This clean covariance block is local to the frozen-attention, edgewise-consensus calculation. Away from consensus the sender Hessian carries additional residual-dependent terms, and envelope elimination adds the covariance correction in Eq.~\eqref{eq:reduced_row_hessian}. Equation~\eqref{eq:mass_covariance} therefore does not assert that the fully reduced covariance Hessian is globally positive definite.

%-----------------------------------------------------------------------
\subsubsection{Mean-Covariance Cross Blocks}
%-----------------------------------------------------------------------

\paragraph{Prior term:}
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| p_i)}{\partial \mu_i \partial \Sigma_i} = 0
\end{equation}

\paragraph{Sensory term:} The sensory free energy decomposes into a component quadratic in $\mu_i$, namely $(o_i - \mu_i)^T\Lambda_{o_i}(o_i - \mu_i)$, and a component linear in $\Sigma_i$, namely $\mathrm{tr}(\Lambda_{o_i}\Sigma_i)$. These are independent, so:
\begin{equation}
[\mathbf{H}_F^{\mu\Sigma}]_{ii}^{\text{sensory}} = 0
\end{equation}

\paragraph{Consensus (cross-agent):} From $\partial D_{\mathrm{KL}}(q_i \| \tilde{q}_k)/\partial \mu_i = \tilde{\Lambda}_{q_k}(\mu_i - \tilde{\mu}_k)$, varying $\Sigma_k$:
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q_i \| \tilde{q}_k)}{\partial \mu_i \partial \Sigma_k}[V] = -\Omega_{ik}^{-T}\Lambda_{q_k}V\Lambda_{q_k}\Omega_{ik}^{-1}(\mu_i - \tilde{\mu}_k)
\end{equation}

This vanishes at consensus ($\mu_i = \tilde{\mu}_k$):
\begin{equation}
[\mathbf{H}_F^{\mu\Sigma}]_{ik} = 0 \quad \text{when } \mu_i = \tilde{\mu}_k
\end{equation}

%-----------------------------------------------------------------------
\subsection{Complete Local Stiffness Assembly}
%-----------------------------------------------------------------------

The full state vector is $\xi = (\mu_1, \ldots, \mu_N, \Sigma_1, \ldots, \Sigma_N)$.

\subsubsection{Block Structure}

\begin{equation}
\mathbf H_F = \begin{pmatrix} \mathbf H_F^\mu & \mathbf H_F^{\mu\Sigma} \\ (\mathbf H_F^{\mu\Sigma})^T & \mathbf H_F^\Sigma \end{pmatrix}
\end{equation}
where each block is an $N \times N$ matrix of sub-blocks.

\subsubsection{Explicit Formulae}

\paragraph{Mean sector diagonal:}
\begin{equation}
[\mathbf H_F^\mu]_{ii} = \underbrace{\alpha_i\bar{\Lambda}_{p_i}}_{\text{prior anchoring}} + \underbrace{\sum_k \beta_{ik}\Omega_{ik}^{-T}\Lambda_{q_k}\Omega_{ik}^{-1}}_{\text{incoming consensus}} + \underbrace{\sum_j \beta_{ji}\Lambda_{q_i}}_{\text{outgoing recoil}} + \underbrace{\Lambda_{o_i}}_{\text{sensory grounding}}
\label{eq:mass_full_diagonal}
\end{equation}

\paragraph{Mean sector off-diagonal:}
\begin{equation}
[\mathbf H_F^\mu]_{ik} = -\beta_{ik}\Omega_{ik}^{-T}\Lambda_{q_k} - \beta_{ki}\Lambda_{q_i}\Omega_{ki}^{-1} \quad (i \neq k)
\end{equation}

\paragraph{Covariance sector diagonal:}
\begin{equation}
[\mathbf H_F^\Sigma]_{ii} = \frac{1}{2}(\Lambda_{q_i} \otimes \Lambda_{q_i}) \cdot \left(\alpha_i + \sum_k \beta_{ik} + \sum_j \beta_{ji}\right)
\end{equation}

\paragraph{Cross mean-covariance (at consensus):}
\begin{equation}
[\mathbf H_F^{\mu\Sigma}]_{ik} = 0 \quad \text{when } \mu_i = \tilde{\mu}_k
\end{equation}

%-----------------------------------------------------------------------
\subsection{Interpretation}
%-----------------------------------------------------------------------

\subsubsection{Four-Part Mean Stiffness}

The mean-sector diagonal stiffness for agent $i$ is
\begin{equation}
\boxed{
[\mathbf H_F^\mu]_{ii} = \alpha_i\bar{\Lambda}_{p_i} + \sum_k \beta_{ik}\tilde{\Lambda}_{q_k} + \sum_j \beta_{ji}\Lambda_{q_i} + \Lambda_{o_i}
}
\label{eq:effective_mass}
\end{equation}

The four summands are prior, incoming relational, outgoing relational/recoil, and sensory stiffness. They describe how the selected scalar objective resists an affine-coordinate displacement; they do not yet specify how much kinetic momentum accompanies a given belief velocity.

\subsubsection{Asymmetry of Sensory Contribution}

The sensory precision $\Lambda_{o_i}$ contributes to the mean-sector stiffness (Eq.~\ref{eq:mass_full_diagonal}) and the mean-sector force (Eq.~\ref{eq:total_gradient_mu}), but not to the covariance-sector stiffness (Eq.~\ref{eq:mass_covariance}). This asymmetry arises because the sensory term is quadratic in $\mu$ but only affine in $\Sigma$.

\subsubsection{Kinetic Energy}

\begin{equation}
T = \frac{1}{2}\dot{\mu}^T\mathbf{M}^\mu\dot{\mu} + \frac{1}{2}\mathrm{tr}\left[\mathbf{M}^\Sigma(\dot{\Sigma}, \dot{\Sigma})\right]
\end{equation}

This display defines a kinetic energy only after a positive $\mathbf M$ has been chosen by the kinetic postulate. The Fisher covariance block is one-half the affine-invariant SPD metric in the usual convention, whereas Eq.~\eqref{eq:mass_covariance} is a loss-Hessian block with objective-dependent coefficients. Their similar Kronecker forms do not identify $\mathbf M$, $\mathbf G$, and $\mathbf H_F$.

%-----------------------------------------------------------------------
\subsection{The Hamiltonian}
%-----------------------------------------------------------------------

After independently choosing a positive kinetic metric $\mathbf M$, define conjugate momenta $\pi=(\pi^\mu,\Pi^\Sigma)=\mathbf M\dot\xi$ and the conditional Hamiltonian
\begin{equation}
\boxed{
H = \frac{1}{2}\langle\pi, \mathbf{M}^{-1}\pi\rangle + \mathcal{F}[\xi]
}
\label{eq:hamiltonian}
\end{equation}

%-----------------------------------------------------------------------
\subsection{Hamilton's Equations}
%-----------------------------------------------------------------------

\subsubsection{Equations of Motion}

\begin{align}
\dot{\mu}_i &= \sum_k [\mathbf{M}^{-1}]_{ik}^{\mu\mu}\pi_k^\mu + \sum_k [\mathbf{M}^{-1}]_{ik}^{\mu\Sigma}\Pi_k^\Sigma \label{eq:hamilton_mu} \\[6pt]
\dot{\Sigma}_i &= \sum_k [\mathbf{M}^{-1}]_{ik}^{\Sigma\mu}\pi_k^\mu + \sum_k [\mathbf{M}^{-1}]_{ik}^{\Sigma\Sigma}\Pi_k^\Sigma \label{eq:hamilton_sigma} \\[6pt]
\dot{\pi}_i^\mu &= -\frac{\partial \mathcal{F}}{\partial \mu_i} - \frac{1}{2}\pi^T\frac{\partial \mathbf{M}^{-1}}{\partial \mu_i}\pi \label{eq:hamilton_pi_mu} \\[6pt]
\dot{\Pi}_i^\Sigma &= -\frac{\partial \mathcal{F}}{\partial \Sigma_i} - \frac{1}{2}\pi^T\frac{\partial \mathbf{M}^{-1}}{\partial \Sigma_i}\pi \label{eq:hamilton_pi_sigma}
\end{align}

\subsubsection{Force Decomposition}

The potential forces decompose into four distinct contributions:
\begin{equation}
\boxed{
-\frac{\partial \mathcal{F}}{\partial \mu_i} = \underbrace{-\alpha_i\bar{\Lambda}_{p_i}(\mu_i - \bar{\mu}_i)}_{\text{prior restoring}} \underbrace{- \sum_k \beta_{ik}\tilde{\Lambda}_{q_k}(\mu_i - \tilde{\mu}_k)}_{\text{consensus}} \underbrace{- \sum_j \beta_{ji}\Lambda_{q_i}\Omega_{ji}^{-1}(\tilde{\mu}_i^{(j)} - \mu_j)}_{\text{reciprocal}} \underbrace{- \Lambda_{o_i}(\mu_i - o_i)}_{\text{sensory evidence}}
}
\label{eq:force_decomposition}
\end{equation}

The metric-derivative force $f_i^{\text{geo}}=-\frac12\sum_{jkl}(\pi_j^\mu)^T\frac{\partial[\mathbf M^{-1}]_{jk}^{\mu\mu}}{\partial\mu_i}\pi_k^\mu$ is present whenever the independently chosen kinetic metric varies with the mean coordinates.  It vanishes in the locally constant kinetic approximation used for the modal formulas.  Freezing attention in the potential does not, by itself, determine whether the separately postulated $\mathbf M$ is state dependent.

\subsubsection{Compact Form}

\begin{equation}
\boxed{
\begin{aligned}
\dot{\xi} &= \mathbf{M}^{-1}\pi \\
\dot{\pi} &= -\nabla\mathcal{F} - \frac{1}{2}\nabla_\xi\langle\pi, \mathbf{M}^{-1}\pi\rangle
\end{aligned}
}
\end{equation}

If the chosen $\mathbf M$ depends on $\Sigma$, the Hamiltonian is nonseparable and requires an integrator that retains the metric-derivative terms.  The local oscillator formulas in the main text instead freeze $\mathbf M$.

%-----------------------------------------------------------------------
\subsection{Damped Dynamics}
%-----------------------------------------------------------------------

With locally constant $\mathbf M$, dissipation is represented by a positive-semidefinite tensor $\boldsymbol\Gamma$:
\begin{equation}
\boxed{
\mathbf M\ddot{\boldsymbol\mu}
+\boldsymbol\Gamma\dot{\boldsymbol\mu}
+\nabla_{\boldsymbol\mu}\mathcal F
=f_{\mathrm{ext}}.}
\label{eq:damped_dynamics}
\end{equation}
Linearization introduces the restoring stiffness $\mathbf K$.  Nontrivial oscillator spectra require $\mathbf K$ to be operationally independent of $\mathbf M$; if both are the same $\mathbf H_F$, Eq.~\eqref{eq:hessian_mass_degeneracy} gives a unit spectrum up to scale.  Only after projection onto a positive modal pair $(m,k)$ with modal damping $\gamma$ does the scalar discriminant $\gamma^2-4mk$ classify over-, critical-, and underdamping.

%-----------------------------------------------------------------------
\subsection{Momentum Current with Sensory Coupling}
%-----------------------------------------------------------------------

For a directed edge $k\to i$, the receiver-frame momentum current is
\begin{equation}
J_{k \to i}=\beta_{ik}\widetilde\Lambda_{q_k}
(\Omega_{ik}\mu_k-\mu_i),
\end{equation}

and the same edge gives the sender recoil $-\Omega_{ik}^\top J_{k\to i}$.  Hence the complete block balance is
\begin{align}
\dot\pi_i^\mu
&+(\boldsymbol\Gamma\dot{\boldsymbol\mu})_i
+\alpha_i\bar\Lambda_{p_i}(\mu_i-\bar\mu_i)
+\Lambda_{o_i}(\mu_i-o_i)\nonumber\\
&=\sum_kJ_{k\to i}
-\sum_j\Omega_{ji}^\top J_{i\to j}
+f_{i,\mathrm{ext}}.
\end{align}
Fixed asymmetric attention still defines the scalar potential whose complete derivative gives this equation.  Dropping the outgoing recoil is a receiver-only detached truncation and is the nonconservative approximation, not a consequence of row normalization itself.

%-----------------------------------------------------------------------
\subsection{Summary}
%-----------------------------------------------------------------------

\begin{tcolorbox}[colback=gray!5,colframe=gray!75,title=The Complete Theory with Sensory Evidence]

\textbf{State:} Each agent $i$ has belief $q_i = \mathcal{N}(\mu_i, \Sigma_i)$ with fixed prior $p_i = \mathcal{N}(\bar{\mu}_i, \bar{\Sigma}_i)$ and observations $o_i$ with precision $\Lambda_{o_i}$.

\textbf{Free Energy:} The unique population scalar is the entropy-retaining $\mathcal F_{\mathrm{full}}$ of Eq.~\eqref{eq:canonical_full_vfe}; fixed attention makes its categorical term constant only for the belief-coordinate derivatives in this appendix.

\textbf{Three geometries:} $\mathbf G$ is the intrinsic Fisher--Rao metric, $\mathbf H_F$ is the affine-coordinate loss stiffness, and $\mathbf M$ is a separately postulated positive kinetic metric.  A positive local restriction of $\mathbf H_F$ may be chosen as $\mathbf M$ only under the kinetic-metric postulate; the objects are not globally identical.

\textbf{Coupled momentum:}
\begin{equation}
\pi_i^\mu=\sum_k\mathbf M_{ik}^\mu\dot\mu_k,
\end{equation}
with $\pi_i^\mu=M_i\dot\mu_i$ reserved for a block-diagonal approximation.

\textbf{Dynamics:}
\begin{equation}
\dot{\xi} = \mathbf{M}^{-1}\pi, \qquad \dot{\pi} = -\nabla\mathcal{F} - \frac{1}{2}\nabla_\xi\langle\pi, \mathbf{M}^{-1}\pi\rangle
\end{equation}

\textbf{Scope:} Equation~\eqref{eq:primary_fisher_flow} remains the primary dynamics.  The Hamiltonian equations in this box apply only after the kinetic postulate and do not turn four-part VFE stiffness into an information-geometric theorem about inertia.

\end{tcolorbox}

%==============================================================================
% APPENDIX: WHY FORWARD KL DIVERGENCE
%==============================================================================

\section{Why Forward KL Divergence?}
\label{app:forward_kl}

A natural question arises: why use the forward KL divergence $D_{\mathrm{KL}}(q_i \| \Omega_{ik}\cdot q_k)$ rather than the reverse $D_{\mathrm{KL}}(\Omega_{ik}\cdot q_k \| q_i)$ in the social coupling term? This appendix provides both intuitive and formal justifications.

\subsection{Intuitive Argument: Zero-Forcing vs.\ Mean-Seeking}

Following the standard convention in variational inference, where $q$ is the optimized argument, the forward KL $D_{\mathrm{KL}}(q \| p)$ is "mode-seeking" or "zero-forcing": it penalizes $q$ for placing probability mass where $p$ has little, forcing $q$ to vanish off the support of $p$ and to concentrate on a mode. The reverse KL $D_{\mathrm{KL}}(p \| q)$ is "mean-seeking" or "mass-covering": it penalizes $q$ for failing to cover regions where $p$ has mass, encouraging $q$ to spread across the support of $p$.

For social influence, the forward direction $D_{\mathrm{KL}}(q_i \| \Omega_{ik}\cdot q_k)$ is the natural one. Agent $i$ is penalized for holding beliefs that the transported neighbor belief considers unlikely: if agent $i$ is certain about something that neighbor $k$ (in $i$'s frame) deems improbable, $i$ pays a high cost. This zero-forcing behavior matches the intuition that an agent's confidence is curbed by disagreement with attended neighbors.

\subsection{Formal Uniqueness Argument}

Consider the general family of $f$-divergences $D_f(q \| p) = \int p(x) f(q(x)/p(x)) dx$ for convex $f$ with $f(1) = 0$. We seek divergences satisfying the following desiderata for social coupling:

\paragraph{Desideratum 1: Convexity in $q$.} The coupling term should be convex in agent $i$'s belief $q_i$, ensuring a well-defined optimization landscape. All $f$-divergences satisfy this when the first argument varies.

\paragraph{Desideratum 2: Additivity under transport.} Under the gauge transport $\Omega_{ik}$, the divergence should decompose sensibly. For Gaussians, this requires the divergence to depend on beliefs only through sufficient statistics $(\mu, \Sigma)$.

\paragraph{Desideratum 3: Fisher information as Hessian.} The Hessian of the coupling term at $q_i = \Omega_{ik}\cdot q_k$ should recover the Fisher information metric, ensuring consistency with the mass interpretation. The forward KL satisfies this: for $D_{\mathrm{KL}}(q \| p)$, we have $\nabla^2_q D_{\mathrm{KL}}(q \| p)\big|_{q=p} = \mathcal{I}_p$, the Fisher information at $p$.

This property does not by itself single out the forward direction: the reverse KL has the same local Fisher quadratic form at matching. We therefore present the forward direction as a modeling choice compatible with the variational convention in the next subsection, not as an unconditional uniqueness result. Appendix~\ref{app:social_coupling} supplies a narrower rowwise identity whose first argument is $q_i$; its frozen-source scope does not strengthen that identity into a population generative-model derivation.

\subsection{Connection to Variational Inference}

In variational inference, the forward KL $D_{\mathrm{KL}}(q \| p)$ is the natural objective when $q$ is the variational approximation and $p$ is the target. Minimizing this drives $q$ toward $p$ in an information-theoretic sense. Our social coupling term $D_{\mathrm{KL}}(q_i \| \Omega_{ik}\cdot q_k)$ follows this convention: agent $i$'s belief $q_i$ is driven toward the transported neighbor belief $\Omega_{ik}\cdot q_k$. The reverse direction would have neighbors' beliefs driving toward agent $i$, inverting the causal structure of social influence.

%==============================================================================
% APPENDIX: DERIVATION OF SOCIAL COUPLING FROM NORMALIZED GENERATIVE MODEL
%==============================================================================

\section{Social Coupling from a Mixture-of-Sources Consensus Energy}
\label{app:social_coupling}

This appendix records the limited probabilistic identity that motivates the engineered consensus row without repeating the authoritative optimization in Eqs.~\eqref{eq:attention_row_objective}--\eqref{eq:reduced_vfe_envelope}. Fix receiver $i$ and define the transported source templates $r_{ij}=\Omega_{ij\#}q_j$. Under the rowwise source-independence assumption, set
\begin{equation}
Q_i(c,j)=q_i(c)\beta_{ij},
\qquad
P_i(c,j)=r_{ij}(c)\pi_{ij}.
\label{eq:row_source_models}
\end{equation}
Both distributions are normalized when $q_i$ and every $r_{ij}$ are normalized, $\sum_j\beta_{ij}=1$, and $\sum_j\pi_{ij}=1$. Direct expansion gives the exact unit-temperature identity
\begin{equation}
\KL(Q_i\|P_i)
=\sum_j\beta_{ij}\KL(q_i\|r_{ij})
+\sum_j\beta_{ij}\log\frac{\beta_{ij}}{\pi_{ij}}.
\label{eq:row_source_kl_identity}
\end{equation}
Thus a fixed row has an ordinary joint-KL interpretation when the categorical coefficient is one. For general $\tau$, Eq.~\eqref{eq:attention_row_objective} is instead a tempered loss-based row objective; multiplying only its categorical KL by $\tau$ does not preserve the single-KL identity in Eq.~\eqref{eq:row_source_kl_identity}.

The scope restriction is essential. The templates $r_{ij}=\Omega_{ij\#}q_j$ are other agents' variational beliefs, not fixed factors of one population generative model. Equation~\eqref{eq:row_source_kl_identity} is therefore exact conditional on frozen row templates, but it does not establish that the sum over interacting rows is the negative ELBO of one fixed joint distribution over the original agent states. Nor does it establish existence, uniqueness, or convergence of a simultaneous population fixed point. This is precisely fence E1 in Table~\ref{tab:epistemic-status}.

The canonical consequences are consequently taken from the scalar objective itself, not from a stronger generative-model claim. The row minimizer is Eq.~\eqref{eq:attention_weights}, the entropy-retaining reduced value is Eq.~\eqref{eq:canonical_row_reduced_vfe}, and its differential is Eq.~\eqref{eq:reduced_vfe_envelope}. The entropy-suppressed $S_i$ in Eq.~\eqref{eq:entropy_suppressed_surrogate} remains a distinct comparator with the covariance response in Eq.~\eqref{eq:entropy_suppressed_differential}.


%==============================================================================
% APPENDIX: GENERALIZATION BEYOND GAUSSIAN BELIEFS
%==============================================================================

\section{Generalization Beyond Gaussian Beliefs}
\label{app:exponential_family}

The Gaussian assumption, while analytically convenient, restricts the theory to unimodal beliefs with elliptical uncertainty. This appendix outlines how the framework extends to exponential family distributions and discusses the challenges of fully general distributions.

\subsection{Exponential Family Formulation}

An exponential family distribution takes the form:
\begin{equation}
p(x \mid \theta) = h(x) \exp\left[\theta^T T(x) - A(\theta)\right]
\end{equation}
where $\theta \in \mathbb{R}^k$ are natural parameters, $T(x)$ are sufficient statistics, and $A(\theta)$ is the log-partition function ensuring normalization.

\subsubsection{Fisher Geometry and Local KL Stiffness}

The Fisher information matrix for exponential families has the elegant form:
\begin{equation}
\mathcal{I}(\theta) = \nabla^2 A(\theta) = \mathrm{Cov}_{p(\cdot|\theta)}[T(x)]
\end{equation}

This is the Hessian of the log-partition function, equivalently the covariance of sufficient statistics. It defines the intrinsic metric $\mathbf G$ of the exponential family. The coordinate Hessian of a KL objective agrees with this metric at the matching point:
\begin{equation}
\boxed{\mathbf H_{\mathrm{KL}}(\theta_i)=\nabla^2 A(\theta_i)=\mathbf G(\theta_i)}
\qquad (\theta_q=\theta_p).
\end{equation}

For a Gaussian with fixed covariance, this gives $\mathbf H_{\mathrm{KL},\mu\mu}=\mathbf G_{\mu\mu}=\Lambda$ at matching. It does not choose a kinetic metric $\mathbf M$.

\subsubsection{KL Divergence Structure}

For exponential families, the KL divergence takes the Bregman divergence form:
\begin{equation}
D_{\mathrm{KL}}(q \| p) = A(\theta_p) - A(\theta_q) - \nabla A(\theta_q)^T(\theta_p - \theta_q)
\end{equation}

Differentiating twice with respect to $\theta_q$ gives the gradient $\nabla^2 A(\theta_q)(\theta_q - \theta_p)$ and the Hessian
\begin{equation}
\frac{\partial^2 D_{\mathrm{KL}}(q \| p)}{\partial \theta_q \partial \theta_q^T} = \nabla^2 A(\theta_q) + \nabla^3 A(\theta_q)\cdot(\theta_q - \theta_p) ,
\end{equation}
where $\nabla^3 A$ is the third-cumulant tensor of the sufficient statistics contracted against $\theta_q-\theta_p$. This Hessian reduces to the Fisher information $\mathcal I(\theta_q)=\nabla^2A(\theta_q)$ only at matching, or globally when $A$ is quadratic. Thus the loss-Hessian/Fisher equality is a local matching or quadratic-regime statement, not a global identity and not an inertial conclusion.

\subsection{Multi-Modal Beliefs: Mixture Models}

Real beliefs are often multi-modal, representing hypothesis competition or attitude ambivalence. Gaussian mixtures provide a tractable extension:
\begin{equation}
q(x) = \sum_{m=1}^M w_m \mathcal{N}(x; \mu_m, \Sigma_m), \quad \sum_m w_m = 1
\end{equation}

\subsubsection{Extended State Space}

The state space becomes $\xi = (\{w_m\}, \{\mu_m\}, \{\Sigma_m\})$, comprising mixture weights on the $(M-1)$-simplex, component means in $\mathbb{R}^{d \times M}$, and component covariances in $\mathrm{SPD}(d)^M$.

\subsubsection{Fisher Information for Mixtures}

The Fisher information for mixture models does not decompose cleanly. The full Fisher matrix couples all parameters:
\begin{equation}
\mathcal{I} = \begin{pmatrix} \mathcal{I}^{ww} & \mathcal{I}^{w\mu} & \mathcal{I}^{w\Sigma} \\ (\mathcal{I}^{w\mu})^T & \mathcal{I}^{\mu\mu} & \mathcal{I}^{\mu\Sigma} \\ (\mathcal{I}^{w\Sigma})^T & (\mathcal{I}^{\mu\Sigma})^T & \mathcal{I}^{\Sigma\Sigma} \end{pmatrix}
\end{equation}

The weight-weight block takes the form:
\begin{equation}
\mathcal{I}^{ww}_{mn} = \int \frac{1}{q(x)} \frac{\partial q}{\partial w_m} \frac{\partial q}{\partial w_n} dx = \int \frac{\mathcal{N}_m(x) \mathcal{N}_n(x)}{q(x)} dx
\end{equation}

This integral generally requires numerical evaluation but captures the intuition that weight-inertia depends on component overlap: well-separated modes have nearly independent weight dynamics.

\subsection{Fully General Distributions}

For arbitrary distributions $q(x)$ without parametric form, the state space becomes infinite-dimensional (a functional space). The Fisher-Rao metric generalizes to:
\begin{equation}
\langle \delta q_1, \delta q_2 \rangle_q = \int \frac{\delta q_1(x)   \delta q_2(x)}{q(x)} dx
\end{equation}

This defines a Riemannian metric on the space of probability densities. Gradient flow with respect to this metric yields the natural gradient dynamics studied in information geometry \citep{Amari2016}.


\subsection{Gauge Structure for Non-Gaussian Beliefs}

The gauge transport operators $\Omega_{ij}$ generalize beyond rotations when beliefs are non-Gaussian.

For exponential families transport acts on natural parameters: $\theta \mapsto \Omega \cdot \theta$ where $\Omega$ preserves the parameter space structure.

In the case of mixtures transport can permute components, rotate within-component means, and transform covariances.

For general distributions transport becomes a diffeomorphism of the sample space such that $q(x) \mapsto q(\Omega^{-1}(x)) |\det D\Omega^{-1}|$.

The gauge group expands from $\mathrm{SO}(d)$ to the group of measure-preserving diffeomorphisms, with corresponding enlargement of the connection structure.

\subsection{Summary}

The intrinsic Fisher metric and its natural-gradient flow generalize to regular exponential families and, with functional-analytic qualifications, to spaces of positive densities.  A kinetic extension does not generalize automatically: each statistical family still requires a separately declared positive kinetic metric, and any identification with a loss-Hessian restriction is a modeling postulate subject to the degeneracy in Eq.~\eqref{eq:hessian_mass_degeneracy}.  Mixtures add singularities and cross-parameter Fisher blocks, so neither the Gaussian kinetic choice nor its oscillator formulas transfer without further analysis.

\bibliographystyle{apalike}
\bibliography{references}

\end{document}
